---
title: "Tandayapa AudioMoth spacing — Day-1 pilot analysis"
subtitle: "ACI, BI, ADI per recorder → pairwise dissimilarity vs distance (forest vs pasture)"
author: "Franz Chandi, Maria Andrade, Priscila Cajas, Noemi Castro"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
    self_contained: false   # keep audio/spectrograms as light external links
    keep_md: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
```

# What this script does

This is a **pilot for the first deployment day**. The goal is to check that the
whole pipeline runs end-to-end so we can later repeat it across all days.

Pipeline (one model *per index*, exactly as in the project plan):

1. Find every WAV, read its start time from the filename, **correct device clock
   offsets**, and keep only clips inside the day-1 analysis window.
2. Compute the three soundscape indices from the research question with
   **`soundecology`** (uses `seewave`/`tuneR` underneath): **ACI**, **BI**, **ADI**.
3. Compute a **mean spectrum per clip** → a **rain/wind QC flag** (low-frequency spikes).
4. Average the clips to **one value per recorder per hour** (clips/hours are
   *temporal subsamples*, never independent replicates).
5. Build the **pairwise table**: for each within-habitat recorder pair and hour,
   `dissimilarity = abs(index_i - index_j)`, attach `distance`, `habitat`, `pair`, `hour`.
6. Add a **direct spectral similarity** (`1 - cor` of the two clips' spectra) as an
   extra, more literal answer to "did they record the same sound?".
7. Plot dissimilarity vs distance (forest vs pasture) and fit a **pilot model** per index.
8. *(Optional)* run **`birdnetR`** to get bird species per clip as a community-level
   cross-check.

The processing is **parallelised** (`furrr`) using ~2/3 of cores so the machine
stays usable; results are cached so re-renders are instant.

> **Phrasing rule kept throughout:** we talk about recordings becoming *more
> different / less similar* with distance. We never say "the indices increase
> with distance".

> **Day-1 caveat:** with a single day we cannot fit a `(1 | day)` random effect
> and we only have two realised distances (15 m and 30 m). The model here is a
> *pilot* `lm`; the multi-day `lmer`/`glmmTMB` versions are written but disabled
> until more days are added.

---

# 0. Install packages (run once)

```{r install, eval=FALSE}
# Core acoustic + data stack
install.packages(c(
  "soundecology",  # ACI, BI, ADI (wraps seewave/tuneR)
  "tuneR",         # read WAV
  "seewave",       # signal helpers
  "lubridate",     # time handling
  "dplyr", "tidyr", "stringr", "purrr", "readr", "ggplot2",
  "lme4", "lmerTest",  # multi-day mixed model (for later)
  "future", "furrr"    # parallel clip processing (Windows-safe)
))

# Bird ID with BirdNET. birdnetR (>= 0.3) auto-provisions a Python backend via
# reticulate the first time you create a model — no separate install step.
# install.packages(c("birdnetR", "reticulate"))
# Then run the standalone runner once:  Rscript run_birdnet_day1.R
# (first model creation downloads TensorFlow + the model, ~10-15 min one time)
```

---

# 1. Parameters — edit these

```{r params}
library(lubridate)
## Devices, windows, spacing, distance and inventory all live in the shared file.
source(file.path("d:/Audiomoths", "tandayapa_common.R"))

## --- Which deployment day this report covers ----------------------------
## Override without editing:  Sys.setenv(TANDAYAPA_DAY="day2") before render.
DAY <- Sys.getenv("TANDAYAPA_DAY", "day1")
stopifnot(DAY %in% names(DAY_WINDOWS))

## --- Per-day output folder + cache --------------------------------------
root_dir  <- ROOT_DIR
out_dir   <- day_dir(DAY)                          # outputs/<DAY>/ (with clips/)
cache_csv <- file.path(out_dir, "indices.csv")

.w <- DAY_WINDOWS[[DAY]]
window_start <- ymd_hms(.w[1], tz = TZ)
window_end   <- ymd_hms(.w[2], tz = TZ)
SPACING_M    <- DAY_SPACING[[DAY]]                 # 15 (day1) / 30 (day2)

## --- Parallel processing -----------------------------------------------
## The per-clip work is independent across clips, so it scales ~linearly with
## cores. We use ~2/3 of cores so the machine stays usable while it runs.
## Flip USE_PARALLEL <- FALSE to debug serially; lower WORKER_FRACTION if it lags.
USE_PARALLEL    <- TRUE
WORKER_FRACTION <- 2/3
N_WORKERS       <- max(1, round(parallel::detectCores() * WORKER_FRACTION))

## --- Which indices to carry through the whole pipeline ------------------
## The three named in the research question. (NDSI/AEI were dropped: AEI is
## ~0.9 redundant with ADI, and NDSI sits outside the RQ.) SPECTRAL similarity
## and BirdNET species-overlap are added later as distinct response families.
INDICES <- c("aci", "bi", "adi")

## --- Index settings (soundecology defaults, exposed so they're explicit) -
SAMPLE_RATE <- 48000                 # from CONFIG.TXT
idx_cfg <- list(
  aci  = list(fft_w = 512, j = 5),
  bi   = list(min_freq = 2000, max_freq = 8000, fft_w = 512),
  adi  = list(max_freq = 10000, db_threshold = -50, freq_step = 1000)
)

## --- Spectral similarity + rain/wind QC settings ------------------------
spectra_rds  <- file.path(out_dir, "spectra.rds")
SPEC_WL      <- 1024     # FFT window for mean spectra (bins = SPEC_WL/2)
LF_MAX_HZ    <- 1500     # "low frequency" ceiling for the rain/wind proxy
## Natural soundscapes are already low-frequency-heavy, so we flag *spikes*
## above each recorder's own baseline, not an absolute level:
## flag if lf_ratio > median + LF_K * MAD (per recorder).
LF_K         <- 3
```

---

# 2. Build the file inventory

We parse the start time from each filename (`YYYYMMDD_HHMMSS.WAV`), add the
per-device clock correction, then keep clips inside the window.

```{r inventory}
library(dplyr); library(stringr); library(tidyr); library(purrr)

## Clock-corrected inventory for this DAY's window (from tandayapa_common.R)
wav_files <- build_inventory(DAY)

## Sanity check: coverage per device (watch for F1503 stopping early)
inventory_summary <- wav_files %>%
  group_by(folder, habitat, point, moth_id, offset_h) %>%
  summarise(n_clips = n(),
            first = min(time), last = max(time), .groups = "drop")
inventory_summary
```

```{r inventory-note, echo=FALSE, results='asis'}
cat(sprintf("\n**Window:** %s → %s (%s).  ",
            format(window_start), format(window_end), TZ))
cat(sprintf("**Total clips in window:** %d.\n", nrow(wav_files)))
```

---

# 3. Compute acoustic indices per clip (ACI, BI, ADI)

`soundecology` prints progress for every file; we silence that and store the three
indices named in the research question: **ACI**, **BI**, **ADI**. Results are **cached**
to `outputs/<DAY>/indices.csv` — delete that file (or add a new index to `INDICES`) to
force a recompute. Failed reads are kept as `NA` so one bad clip can't stop the run.

```{r compute-indices}
library(soundecology); library(tuneR); library(readr)

one_clip_indices <- function(path) {
  out <- list(aci = NA_real_, bi = NA_real_, adi = NA_real_)
  res <- try({
    w <- readWave(path)
    invisible(utils::capture.output({
      aci  <- acoustic_complexity(w, fft_w = idx_cfg$aci$fft_w, j = idx_cfg$aci$j)
      bi   <- bioacoustic_index(w, min_freq = idx_cfg$bi$min_freq,
                                max_freq = idx_cfg$bi$max_freq, fft_w = idx_cfg$bi$fft_w)
      adi  <- acoustic_diversity(w, max_freq = idx_cfg$adi$max_freq,
                                 db_threshold = idx_cfg$adi$db_threshold,
                                 freq_step = idx_cfg$adi$freq_step)
    }))
    out$aci  <- aci$AciTotAll_left
    out$bi   <- bi$left_area
    out$adi  <- adi$adi_left
  }, silent = TRUE)
  if (inherits(res, "try-error")) message("Failed: ", basename(path))
  as_tibble(out)
}

## Use cache only if it already contains every index in INDICES
cache_ok <- file.exists(cache_csv) &&
  all(INDICES %in% names(read_csv(cache_csv, n_max = 0, show_col_types = FALSE)))

if (cache_ok) {
  indices <- read_csv(cache_csv, show_col_types = FALSE) %>%
    mutate(time = with_tz(time, TZ), hour = with_tz(hour, TZ))
  message("Loaded cached indices: ", nrow(indices), " clips")
} else {
  t0 <- Sys.time()
  if (USE_PARALLEL && requireNamespace("furrr", quietly = TRUE)) {
    library(furrr); library(future)
    plan(multisession, workers = N_WORKERS)     # Windows-safe (PSOCK)
    on.exit(plan(sequential), add = TRUE)       # always release workers
    idx_vals <- future_map_dfr(
      wav_files$paths, one_clip_indices,
      .options = furrr_options(seed = TRUE,
                               packages = c("soundecology", "tuneR", "tibble"),
                               globals  = c("one_clip_indices", "idx_cfg")),
      .progress = TRUE)
    message("Computed in parallel on ", N_WORKERS, " workers")
  } else {
    idx_vals <- map_dfr(wav_files$paths, one_clip_indices)  # serial fallback
  }
  indices <- bind_cols(
    wav_files %>% select(folder, habitat, point, moth_id, paths, time, hour, day),
    idx_vals
  )
  write_csv(indices, cache_csv)
  message(sprintf("Computed + cached %d clips in %.1f min",
                  nrow(indices), as.numeric(difftime(Sys.time(), t0, units = "mins"))))
}

indices %>% select(folder, time, all_of(INDICES)) %>% head(10)
```

```{r index-qc}
## Quick QC: how many clips, how many failed per device
indices %>%
  group_by(folder) %>%
  summarise(n = n(),
            n_failed = sum(dplyr::if_all(all_of(INDICES), is.na)),
            .groups = "drop")
```

---

# 3b. Mean spectra → rain/wind QC flag

Indices react to rain and wind, so we compute a **mean power spectrum per clip**
(reused below for spectral similarity) and a simple proxy: the share of spectral
amplitude **below `LF_MAX_HZ`**. Natural soundscapes are already low-frequency-heavy,
so we flag **spikes above each recorder's own baseline** (`median + LF_K·MAD`), which
is what wind/rain bouts look like — a *flag for review*, not an automatic exclusion.
Cached to `outputs/<DAY>/spectra.rds`.

```{r spectra-compute}
library(seewave)

spec_one <- function(path, wl = SPEC_WL, lf_max_hz = LF_MAX_HZ) {
  out <- list(amp = NULL, rms = NA_real_, lf_ratio = NA_real_, ok = FALSE)
  res <- try({
    w  <- readWave(path)
    x  <- as.numeric(w@left) / (2^(w@bit - 1))          # ints -> ~[-1,1]
    ms <- meanspec(w, f = w@samp.rate, wl = wl, norm = TRUE, plot = FALSE)
    amp <- ms[, 2]; freq_hz <- ms[, 1] * 1000           # seewave returns kHz
    out$amp <- amp
    out$rms <- sqrt(mean(x^2))
    out$lf_ratio <- sum(amp[freq_hz < lf_max_hz]) / sum(amp)
    out$ok  <- TRUE
  }, silent = TRUE)
  out
}

if (file.exists(spectra_rds)) {
  spec_cache <- readRDS(spectra_rds)
  message("Loaded cached spectra")
} else {
  t0 <- Sys.time()
  if (USE_PARALLEL && requireNamespace("furrr", quietly = TRUE)) {
    library(furrr); library(future)
    plan(multisession, workers = N_WORKERS); on.exit(plan(sequential), add = TRUE)
    spec_list <- future_map(wav_files$paths, spec_one,
      .options = furrr_options(seed = TRUE, packages = c("tuneR", "seewave"),
                               globals = c("spec_one", "SPEC_WL", "LF_MAX_HZ")))
  } else {
    spec_list <- map(wav_files$paths, spec_one)
  }
  ok_idx <- vapply(spec_list, `[[`, logical(1), "ok")
  nbin   <- length(spec_list[[which(ok_idx)[1]]]$amp)
  S      <- do.call(rbind, lapply(spec_list,
              function(z) if (isTRUE(z$ok)) z$amp else rep(NA_real_, nbin)))
  f1     <- meanspec(readWave(wav_files$paths[which(ok_idx)[1]]),
                     f = SAMPLE_RATE, wl = SPEC_WL, norm = TRUE, plot = FALSE)
  spec_cache <- list(
    S = S, freq_khz = f1[, 1],
    meta = wav_files %>%
      select(folder, habitat, point, moth_id, time, hour, day) %>%
      mutate(row_id   = row_number(),
             slot     = floor_date(time, "20 minutes"),
             rms      = vapply(spec_list, `[[`, numeric(1), "rms"),
             lf_ratio = vapply(spec_list, `[[`, numeric(1), "lf_ratio")))
  saveRDS(spec_cache, spectra_rds)
  message(sprintf("Computed spectra for %d clips in %.1f min",
                  nrow(spec_cache$meta), as.numeric(difftime(Sys.time(), t0, units = "mins"))))
}

qc <- spec_cache$meta %>%
  group_by(folder) %>%
  mutate(lf_thr = median(lf_ratio, na.rm = TRUE) + LF_K * mad(lf_ratio, na.rm = TRUE),
         flag_rain_wind = lf_ratio > lf_thr) %>%
  ungroup()
```

```{r qc-summary}
qc %>% group_by(habitat, point, moth_id) %>%
  summarise(n = n(), n_flagged = sum(flag_rain_wind),
            pct_flagged = round(100 * mean(flag_rain_wind), 1),
            median_lf = round(median(lf_ratio), 2), .groups = "drop")
```

```{r qc-plot, fig.height=6}
library(ggplot2)
ggplot(qc, aes(time, lf_ratio,
               colour = paste0(toupper(substr(habitat, 1, 1)), point))) +
  geom_line(alpha = 0.7) +
  geom_point(aes(shape = flag_rain_wind, size = flag_rain_wind)) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 4), guide = "none") +
  scale_size_manual(values = c(`FALSE` = 0.6, `TRUE` = 2.4), guide = "none") +
  facet_wrap(~ habitat, ncol = 1) +
  labs(title = "Low-frequency amplitude share over the day (rain/wind proxy)",
       subtitle = paste0("X marks = spikes above the recorder's own baseline (median + ",
                         LF_K, "*MAD) to review for wind/rain."),
       x = NULL, y = paste0("share of amplitude < ", LF_MAX_HZ, " Hz"),
       colour = "Recorder") +
  theme_bw()
```

> To exclude flagged clips from the indices, filter `indices` to the clip times
> where `qc$flag_rain_wind` is `FALSE` before the hourly step. Left off by default
> so day 1 keeps every clip; revisit once you've eyeballed the flagged spectrograms.

---

# 4. Hourly summary (one value per recorder per hour)

```{r hourly}
hourly <- indices %>%
  group_by(habitat, point, moth_id, folder, hour) %>%
  summarise(across(all_of(INDICES), ~ mean(.x, na.rm = TRUE)),
            n_clips = sum(!is.na(aci)),
            .groups = "drop") %>%
  mutate(day = as_date(hour))

hourly %>% arrange(habitat, point, hour) %>% head(12)
```

```{r hourly-plot, fig.height=9}
library(ggplot2); library(tidyr)

hourly_long <- hourly %>%
  pivot_longer(all_of(INDICES), names_to = "index", values_to = "value") %>%
  mutate(index = toupper(index),
         recorder = paste0(toupper(substr(habitat, 1, 1)), point, " (", moth_id, ")"))

ggplot(hourly_long, aes(hour, value, colour = recorder)) +
  geom_line() + geom_point(size = 0.8) +
  facet_grid(index ~ habitat, scales = "free_y") +
  labs(title = "Hourly index values per recorder (day 1)",
       x = NULL, y = "Index value", colour = "Recorder") +
  theme_bw() + theme(legend.position = "bottom")
```

---

# 5. Pairwise dissimilarity vs distance

For each habitat, every within-habitat pair, every hour both recorders are present:
`dissimilarity = abs(index_i - index_j)`.

```{r pairwise}
# All within-habitat point pairs (1-2, 1-3, 2-3)
make_pairs <- function(pts) {
  cb <- combn(sort(pts), 2)
  tibble(point_i = cb[1, ], point_j = cb[2, ])
}
pairs_tbl <- DEVICES %>% filter(deploy == DAY) %>%
  group_by(habitat) %>%
  reframe(make_pairs(point)) %>%
  mutate(distance = point_distance_m(DAY, point_i, point_j),
         pair = paste0(toupper(substr(habitat, 1, 1)), point_i, point_j))

make_pairwise <- function(idx_name) {
  wide <- hourly %>%
    select(habitat, point, hour, value = all_of(idx_name)) %>%
    pivot_wider(names_from = point, values_from = value, names_prefix = "p")
  if (!"p3" %in% names(wide)) wide$p3 <- NA_real_   # day2 pasture has only 2 points
  pairs_tbl %>%
    left_join(wide, by = "habitat", relationship = "many-to-many") %>%
    mutate(
      vi = dplyr::case_when(point_i == 1 ~ p1, point_i == 2 ~ p2, point_i == 3 ~ p3),
      vj = dplyr::case_when(point_j == 1 ~ p1, point_j == 2 ~ p2, point_j == 3 ~ p3),
      dissimilarity = abs(vi - vj),
      index = toupper(idx_name)
    ) %>%
    filter(!is.na(dissimilarity)) %>%
    select(index, habitat, pair, distance, hour, vi, vj, dissimilarity) %>%
    mutate(day = as_date(hour))
}

pairwise <- bind_rows(lapply(INDICES, make_pairwise))
pairwise %>% arrange(index, habitat, pair, hour) %>% head(12)
```

## 5b. Direct spectral similarity (a more literal "same sound?")

`abs(index_i - index_j)` can be near zero even when two recorders hear totally
different sound. So we also compare the **actual mean spectra** of clips recorded
in the **same 20-min slot**: `dissimilarity = 1 - cor(spectrum_i, spectrum_j)`
(0 = identical spectra, → 2 = opposite). Averaged to the hour and appended to the
`pairwise` table as a new "index" called **`SPECTRAL`**, so it flows through every
plot and model below.

```{r spectral-similarity}
S   <- spec_cache$S
key <- spec_cache$meta %>% select(habitat, point, slot, row_id)

spec_pw <- pairs_tbl %>%
  inner_join(key, by = c("habitat", "point_i" = "point"),
             relationship = "many-to-many") %>% rename(row_i = row_id) %>%
  inner_join(key, by = c("habitat", "point_j" = "point", "slot"),
             relationship = "many-to-many") %>% rename(row_j = row_id) %>%
  mutate(r = purrr::map2_dbl(row_i, row_j,
                             ~ suppressWarnings(stats::cor(S[.x, ], S[.y, ]))),
         dissimilarity = 1 - r) %>%
  filter(is.finite(dissimilarity)) %>%
  mutate(hour = floor_date(slot, "hour"))

spec_hourly <- spec_pw %>%
  group_by(habitat, pair, distance, hour) %>%
  summarise(dissimilarity = mean(dissimilarity), .groups = "drop") %>%
  mutate(index = "SPECTRAL", day = as_date(hour))

# Append to the main pairwise table and (re)write the combined CSV
pairwise <- bind_rows(pairwise, spec_hourly)
write_csv(pairwise, file.path(out_dir, "pairwise_dissimilarity.csv"))

spec_hourly %>%
  group_by(habitat, distance) %>%
  summarise(mean_spec_dissim = round(mean(dissimilarity), 3),
            n_hours = n(), .groups = "drop")
```

```{r pairwise-summary}
pairwise %>%
  group_by(index, habitat, distance) %>%
  summarise(mean_dissim = mean(dissimilarity), sd = sd(dissimilarity),
            n_hours = n(), .groups = "drop") %>%
  arrange(index, habitat, distance)
```

```{r pairwise-plot, fig.height=7}
ggplot(pairwise, aes(factor(distance), dissimilarity, fill = habitat)) +
  geom_boxplot(outlier.size = 0.6, alpha = 0.8) +
  facet_wrap(~ index, scales = "free_y") +
  scale_fill_manual(values = c(forest = "#2C5F2D", pasture = "#D9A441")) +
  labs(title = "Pairwise dissimilarity vs recorder distance (day 1)",
       subtitle = "Higher = recordings more different. SPECTRAL = 1-cor(spectra); others = abs(index diff). Only 15 & 30 m on day 1.",
       x = "Distance between recorders (m)", y = "Dissimilarity (free y per panel)") +
  theme_bw()
```

```{r pairwise-diel, fig.height=7}
## Diel view: does the distance gap widen during dawn/dusk choruses? (P3)
ggplot(pairwise, aes(hour, dissimilarity, colour = factor(distance))) +
  geom_line() + geom_point(size = 0.7) +
  facet_grid(index ~ habitat, scales = "free_y") +
  labs(title = "Dissimilarity through the day (day 1)",
       x = NULL, y = "Dissimilarity", colour = "Distance (m)") +
  theme_bw() + theme(legend.position = "bottom")
```

---

# 6. Pilot model

With **one day** we can't fit `(1 | day)`, and distance has only two levels, so
the proper mixed model is overkill / won't identify. For the pilot we fit a simple
`lm` per index to confirm the machinery and read the **distance × habitat** sign.
The multi-day model is written below but disabled.

```{r pilot-lm}
fit_pilot <- function(idx_name) {
  d <- pairwise %>% filter(index == idx_name) %>%
    mutate(distance_z = as.numeric(scale(distance)))
  m <- lm(dissimilarity ~ distance_z * habitat, data = d)
  cat("\n=====", idx_name, "=====\n"); print(summary(m)$coefficients)
  invisible(m)
}
pilot_models <- lapply(sort(unique(pairwise$index)), fit_pilot)
```

```{r full-model, eval=FALSE}
## ----- MULTI-DAY MODEL (enable once several days are processed) -----------
## Stack the per-day `pairwise` tables into one `pairwise_all`, then:
library(lme4); library(lmerTest)

run_lmer <- function(d) {
  d$distance_z <- scale(d$distance)
  m <- lmer(dissimilarity ~ distance_z * habitat + (1 | pair) + (1 | day), data = d)
  print(summary(m)); m
}
# models <- lapply(split(pairwise_all, pairwise_all$index), run_lmer)
## The distance_z:habitatpasture interaction is the key test for P2
## (steeper distance decay in forest than pasture).

## If dissimilarity is bounded/skewed, swap to glmmTMB Gamma/Beta:
# library(glmmTMB)
# glmmTMB(dissimilarity ~ distance_z * habitat + (1|pair) + (1|day),
#         family = Gamma(link = "log"), data = d)
```

---

# 7. BirdNET bird-species cross-check

A community-level complement to the indices: **BirdNET** identifies bird species in
each clip, so we can ask *which species* each recorder hears, not just how the
soundscape "looks". Detections are produced by the standalone, cached runner
**`run_birdnet_day1.R`** (heavy Python/TensorFlow step) → `outputs/<DAY>/birdnet_detections.csv`;
this section just reads that file. If the cache is missing, run the scripts once
(each reads its own `DAY` at the top):

```text
Rscript run_birdnet_day1.R         # ~35 min, scores every clip (confidence 0.25)
Rscript birdnet_location_filter.R  # ~1 min, Tandayapa species list (range filter)
```

A **range filter** then keeps only species BirdNET expects at this location and
time of year, removing out-of-range false positives (e.g. Caribbean/North-American
species the global model can confuse for local ones).

```{r birdnet-load}
det_csv <- file.path(out_dir, "birdnet_detections.csv")
have_birdnet <- file.exists(det_csv)
if (have_birdnet) {
  ## BirdNET returns one row per 3-s segment; segments with nothing above the
  ## confidence threshold come back as NA -> drop them to keep only real detections.
  det <- readr::read_csv(det_csv, show_col_types = FALSE) %>%
    mutate(time = with_tz(time, TZ), hour = with_tz(hour, TZ)) %>%
    filter(!is.na(scientific_name), !is.na(confidence))
  n_sp_raw <- dplyr::n_distinct(det$scientific_name)

  ## Range filter: keep only species BirdNET expects at Tandayapa in early June
  ## (cuts out-of-range false positives). List built by birdnet_location_filter.R.
  loc_csv <- file.path(out_dir, "birdnet_location_species.csv")
  if (file.exists(loc_csv)) {
    allowed <- readr::read_csv(loc_csv, show_col_types = FALSE) %>%
      mutate(scientific_name = sub("_.*$", "", label))
    det <- det %>% filter(scientific_name %in% allowed$scientific_name)
    message("Range filter ON: kept ", dplyr::n_distinct(det$scientific_name),
            " of ", n_sp_raw, " species (", nrow(allowed), " expected locally)")
  } else {
    message("No location list - run birdnet_location_filter.R for the range filter. ",
            "Using all ", n_sp_raw, " species.")
  }
  message("BirdNET detections used: ", nrow(det), " rows")
} else {
  message("No BirdNET cache yet — run run_birdnet_day1.R. Skipping section 7.")
}
```

```{r birdnet-richness, eval=have_birdnet}
## Species set per recorder per hour (a 3-s window may fire many times -> unique)
sp_hour <- det %>%
  group_by(habitat, point, moth_id, hour) %>%
  summarise(species = list(sort(unique(scientific_name))),
            n_detections = n(), .groups = "drop")

## Whole-day richness + activity per recorder
sp_hour %>%
  group_by(habitat, point, moth_id) %>%
  summarise(species_richness = length(unique(unlist(species))),
            total_detections = sum(n_detections),
            active_hours = n(), .groups = "drop") %>%
  arrange(habitat, point)
```

```{r birdnet-top, eval=have_birdnet}
## Most-detected species per habitat (sanity check the IDs make ecological sense)
det %>%
  group_by(habitat, scientific_name, common_name) %>%
  summarise(detections = n(), recorders = n_distinct(point),
            mean_conf = round(mean(confidence), 2), .groups = "drop") %>%
  group_by(habitat) %>% slice_max(detections, n = 8) %>% ungroup() %>%
  arrange(habitat, desc(detections))
```

```{r birdnet-richness-plot, eval=have_birdnet, fig.height=5}
library(ggplot2)
sp_hour %>%
  mutate(n_species = lengths(species),
         recorder = paste0(toupper(substr(habitat, 1, 1)), point)) %>%
  ggplot(aes(hour, n_species, colour = recorder)) +
  geom_line() + geom_point(size = 0.8) +
  facet_wrap(~ habitat, ncol = 1) +
  labs(title = "BirdNET species richness per hour (day 1)",
       x = NULL, y = "Species detected", colour = "Recorder") +
  theme_bw()
```

## 7b. Species-overlap vs distance (independent check on the index story)

For each within-habitat pair and hour, compare the two recorders' species sets with
the **Jaccard index** (shared / total). `1 - Jaccard` is a community dissimilarity:
if recorders far apart sample different birds, it should be **higher at 30 m than 15 m** —
an interpretable, taxonomy-based echo of the acoustic-index and spectral results.

```{r birdnet-jaccard, eval=have_birdnet}
## Presence grid = every recorder-hour that actually recorded (from `hourly`);
## a recorder-hour with no birds gets an empty species set (not dropped).
grid <- hourly %>% select(habitat, point, hour)
sets <- grid %>%
  left_join(sp_hour %>% select(habitat, point, hour, species),
            by = c("habitat", "point", "hour")) %>%
  mutate(species = ifelse(map_lgl(species, is.null), list(character(0)), species))

jaccard <- function(a, b) {
  a <- unique(a); b <- unique(b)
  if (length(a) == 0 && length(b) == 0) return(NA_real_)   # both silent: undefined
  length(intersect(a, b)) / length(union(a, b))
}

wide_sets <- sets %>%
  pivot_wider(names_from = point, values_from = species, names_prefix = "p")

bird_pw <- pairs_tbl %>%
  left_join(wide_sets, by = "habitat", relationship = "many-to-many") %>%
  mutate(
    si = dplyr::case_when(point_i == 1 ~ p1, point_i == 2 ~ p2, point_i == 3 ~ p3),
    sj = dplyr::case_when(point_j == 1 ~ p1, point_j == 2 ~ p2, point_j == 3 ~ p3),
    jac = purrr::map2_dbl(si, sj, jaccard),
    dissimilarity = 1 - jac
  ) %>%
  filter(!is.na(dissimilarity)) %>%
  select(habitat, pair, distance, hour, jaccard = jac, dissimilarity)

readr::write_csv(bird_pw, file.path(out_dir, "birdnet_pairwise.csv"))

bird_pw %>%
  group_by(habitat, distance) %>%
  summarise(mean_jaccard = round(mean(jaccard), 3),
            mean_dissim  = round(mean(dissimilarity), 3),
            n_hours = n(), .groups = "drop")
```

```{r birdnet-jaccard-plot, eval=have_birdnet, fig.height=4.5}
ggplot(bird_pw, aes(factor(distance), dissimilarity, fill = habitat)) +
  geom_boxplot(outlier.size = 0.6, alpha = 0.85) +
  scale_fill_manual(values = c(forest = "#2C5F2D", pasture = "#D9A441")) +
  labs(title = "Bird community dissimilarity (1 - Jaccard) vs distance",
       subtitle = "Higher = recorders hear different species. Day-1 pilot, 15 & 30 m only.",
       x = "Distance between recorders (m)", y = "1 - Jaccard (species sets)") +
  theme_bw()
```

```{r birdnet-jaccard-lm, eval=have_birdnet}
bd <- bird_pw %>% mutate(distance_z = as.numeric(scale(distance)))
summary(lm(dissimilarity ~ distance_z * habitat, data = bd))$coefficients
```

---

# 7c. Validation gallery — spectrograms + audio

For the **top-confidence detections**, the gallery below shows a spectrogram
(x = time, y = frequency) with each BirdNET detection drawn as a **labeled box**
and a **play button** to confirm the call by ear. Built by the standalone
`build_gallery_report.R` (re-scores the chosen clips to recover detection times,
writes PNG + WAV to `outputs/<DAY>/clips/`, and a manifest read here).

> **Box meaning:** BirdNET classifies 3-second segments across the *whole*
> frequency band, so each box is exact in **time** (the detected interval) and
> labelled with the species, but it spans the frequency axis — it is **not** a
> tight frequency box around the song. Audio is linked from `outputs/<DAY>/clips/`, so
> keep the HTML and that folder together for playback.

```{r gallery, results='asis', echo=FALSE}
gm_csv <- file.path(out_dir, "gallery_manifest.csv")
if (file.exists(gm_csv)) {
  gm <- readr::read_csv(gm_csv, show_col_types = FALSE)
  for (i in seq_len(nrow(gm))) {
    g <- gm[i, ]
    cat(sprintf('\n#### %d. %s — *%s* (%s) · confidence %.2f · %s\n\n',
                i, g$recorder, g$common, g$species, g$confidence,
                format(g$time, "%Y-%m-%d %H:%M")))
    cat(sprintf('<img src="%s" alt="%s" style="max-width:100%%;border:1px solid #ccc"/>\n\n',
                g$png, g$id))
    cat(sprintf('<audio controls preload="none" src="%s">your browser cannot play this clip</audio>\n\n',
                g$wav))
  }
} else {
  cat("\n*No gallery yet — run `Rscript build_gallery_report.R` after BirdNET finishes.*\n")
}
```

---

# 8. Outputs & session info

```{r outputs, results='asis'}
cat("Files written to `", out_dir, "`:\n\n", sep = "")
cat("- `indices.csv` — per-clip ACI/BI/ADI (cache)\n")
cat("- `pairwise_dissimilarity.csv` — pairwise table for modelling\n")
```

```{r session}
sessionInfo()
```
