---
title: "Tandayapa AudioMoth spacing — combined analysis (day 1 + day 2)"
subtitle: "Distance decay of recording similarity across 15 / 30 / 60 m, forest vs pasture"
author: "Franz Chandi, Maria Andrade, Priscila Cajas, Noemi Castro"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
    self_contained: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
source(file.path("d:/Audiomoths", "tandayapa_common.R"))
library(readr); library(ggplot2)
HAB_COLS <- c(forest = "#2C5F2D", pasture = "#D9A441")
DAYS <- c("day1", "day2")
```

# What this pools

Each day is analysed on its own (per-day reports + `outputs/dayN/`). This document
**pools both deployments** so distance becomes a real gradient:

- **day1** — 15 m spacing → pairwise distances **15 & 30 m** (forest + pasture).
- **day2** — 30 m spacing → pairwise distances **30 & 60 m** (forest; pasture only
  one pair at 30 m, since P3003 failed).

So across the study we have **15 / 30 / 60 m**. The 30 m point appears in *both*
days, which helps separate a distance effect from a day effect.

> **Read this first — distance is confounded with day.** 15 m occurs *only* in
> day 1 and 60 m *only* in day 2, and the two days have different baseline
> dissimilarity (day 2 is quieter/more similar overall). So a naive 15→30→60 curve
> of the **acoustic indices** can even point the "wrong" way — that's the day
> baseline talking, not distance. Only **30 m bridges both days**. The trustworthy
> evidence is therefore: (1) the **within-day slopes** (§1b), and (2) **BirdNET
> species overlap** (§3), which declines with distance despite the confound. Plots
> below distinguish the two days by point shape so the structure is visible.
>
> Other caveats: only forest reaches 60 m (pasture lost a recorder on day 2); still
> just two days. Treat everything as indicative, not definitive — this is why real
> **GPS distances within a day** (multiple spacings simultaneously) would be the big
> upgrade.

```{r load}
read_day <- function(dy, file) {   # `dy` not `day`: the CSV has a `day` (date) column
  f <- file.path(day_dir(dy), file)
  if (!file.exists(f)) return(NULL)
  read_csv(f, show_col_types = FALSE) %>% mutate(deploy = dy,
                                                 pair_uid = paste(dy, pair))
}
pairwise <- bind_rows(lapply(DAYS, read_day, file = "pairwise_dissimilarity.csv"))
bird     <- bind_rows(lapply(DAYS, function(d) {
  f <- file.path(day_dir(d), "birdnet_pairwise.csv")
  if (!file.exists(f)) return(NULL)
  read_csv(f, show_col_types = FALSE) %>% mutate(deploy = d, pair_uid = paste(d, pair))
}))
cat("Pooled pairwise rows:", nrow(pairwise),
    "| indices:", paste(sort(unique(pairwise$index)), collapse = ", "), "\n")
```

---

# 1. Distance decay (acoustic indices + spectral)

```{r summary}
pairwise %>% group_by(index, habitat, distance) %>%
  summarise(mean_dissim = round(mean(dissimilarity), 3), n_hours = n(), .groups = "drop") %>%
  arrange(index, habitat, distance)
```

Lines connect points **within a day** (day 1: 15→30 m; day 2: 30→60 m) — they are
*not* joined across days, because the day baseline shifts. Point shape = day.

```{r decay-plot, fig.height=7}
ggplot(pairwise, aes(distance, dissimilarity, colour = habitat, linetype = deploy,
                     group = interaction(habitat, deploy))) +
  stat_summary(fun = mean, geom = "line", linewidth = 0.9) +
  stat_summary(fun = mean, geom = "point", size = 2.6) +
  facet_wrap(~ index, scales = "free_y") +
  scale_colour_manual(values = HAB_COLS) +
  scale_x_continuous(breaks = c(15, 30, 60)) +
  labs(title = "Dissimilarity vs distance, within each day (15/30 m = day1, 30/60 m = day2)",
       subtitle = "Lines join within-day points only. Mind the day baseline shift; 30 m is the only shared distance.",
       x = "Distance between recorders (m)", y = "Dissimilarity (free y per panel)", linetype = "Day") +
  theme_bw()
```

## 1b. Within-day slopes (the cleaner read)

Each day on its own, does dissimilarity rise with distance? (Forest spans both
spacings each day; pasture only at the short distance in day 2.)

```{r within-day}
pairwise %>%
  group_by(deploy, index, habitat, distance) %>%
  summarise(mean_dissim = round(mean(dissimilarity), 3), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = distance, values_from = mean_dissim, names_prefix = "m") %>%
  arrange(index, deploy, habitat)
```

---

# 2. Models (pooled)

`lm` shows the distance × habitat picture; the mixed model adds random intercepts
for pair and deployment-day. With two days the `(1|deploy)` term is provisional.

```{r models}
library(lme4); library(lmerTest)
fit_one <- function(idx) {
  d <- pairwise %>% filter(index == idx) %>% mutate(distance_z = as.numeric(scale(distance)))
  cat("\n=====", idx, "=====\n")
  print(round(summary(lm(dissimilarity ~ distance_z * habitat, data = d))$coefficients, 4))
  m <- try(suppressMessages(
    lmer(dissimilarity ~ distance_z * habitat + (1|pair_uid) + (1|deploy), data = d)),
    silent = TRUE)
  if (!inherits(m, "try-error")) {
    co <- summary(m)$coefficients
    if ("distance_z:habitatpasture" %in% rownames(co))
      cat("[lmer interaction t =", round(co["distance_z:habitatpasture","t value"], 2), "]\n")
  }
}
invisible(lapply(sort(unique(pairwise$index)), fit_one))
```

---

# 3. BirdNET community overlap vs distance

```{r bird, eval=!is.null(bird) && nrow(bird) > 0}
bird %>% group_by(habitat, distance) %>%
  summarise(mean_jaccard = round(mean(jaccard), 3),
            mean_dissim  = round(mean(dissimilarity), 3), n_hours = n(), .groups = "drop") %>%
  arrange(habitat, distance)
```

```{r bird-plot, eval=!is.null(bird) && nrow(bird) > 0, fig.height=4.5}
ggplot(bird, aes(distance, dissimilarity, colour = habitat, linetype = deploy,
                 group = interaction(habitat, deploy))) +
  stat_summary(fun = mean, geom = "line", linewidth = 0.9) +
  stat_summary(fun = mean, geom = "point", size = 2.6) +
  scale_colour_manual(values = HAB_COLS) +
  scale_x_continuous(breaks = c(15, 30, 60)) +
  labs(title = "Bird community dissimilarity (1 - Jaccard) vs distance",
       subtitle = "Higher = recorders hear different species. Even across days, overlap falls with distance.",
       x = "Distance between recorders (m)", y = "1 - Jaccard", linetype = "Day") +
  theme_bw()
```

---

# 4. Session info

```{r session}
sessionInfo()
```
