---
title: "Acoustic edge-gradient: how recorder spacing shapes the soundscape in forest and pasture at Tandayapa"
subtitle: "Tandayapa Cloud Forest Station, Ecuador · AudioMoth passive acoustic monitoring · reproducible analysis"
author: "Franz Chandi, Maria Andrade, Priscila Cajas, Noemi Castro"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    fig_width: 6.6
    fig_height: 4.4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
suppressMessages({
  library(tidyverse)   # dplyr, tidyr, readr, purrr, stringr, ggplot2 — cleaning + plots
  library(vegan)       # community dissimilarities (Bray–Curtis, Jaccard)
  library(lme4); library(lmerTest)   # linear mixed models (deploy as random effect)
  library(knitr)
})
CONF_MIN <- 0.5            # primary BirdNET confidence threshold (sensitivity to 0.25 / 0.7 below)
MAXDIST  <- 60            # analyse pairs up to 60 m (drop the lone, unreplicated 120 m pair)
cols <- c(forest = "#2C5F2D", pasture = "#D9A441")

# ---- All analyses read the COMMITTED processed data, so this report knits without the raw audio ----
PROC <- "../data/processed"
det  <- read_csv(file.path(PROC, "birdnet_detections.csv"), show_col_types = FALSE)
idx  <- read_csv(file.path(PROC, "acoustic_indices_per_recorder.csv"), show_col_types = FALSE)
idxd <- read_csv(file.path(PROC, "acoustic_indices_by_diel.csv"), show_col_types = FALSE)
clim <- read_csv(file.path(PROC, "datalogger_microclimate.csv"), show_col_types = FALSE)
day_spacing <- det %>% distinct(deploy, spacing_m) %>% deframe()
```

# Introduction

Passive acoustic monitoring (PAM) with low-cost recorders such as AudioMoth is now a standard tool
for surveying soniferous biodiversity — birds, frogs, and insects — over long periods with minimal
disturbance (Ross et al., 2023). A practical question every PAM study must answer, but rarely tests
directly, is **how far apart recorders should be placed**. If two recorders sit close together they
"hear" almost the same animals and act as one sample (pseudoreplication); if they are far apart they
capture independent slices of the community and provide genuine spatial replication. The right
spacing depends on how fast the recorded soundscape changes with distance, which should depend on the
habitat, because vegetation and terrain control how far sound travels (Bradfer-Lawrence et al., 2023).

Tandayapa spans a mosaic of dense forest and open pasture — an ideal contrast for asking whether
recorder spacing behaves differently in a structurally complex habitat versus an open one. We answer
the question in two currencies and compare them: (1) the **species community** detected by an
automated classifier (BirdNET), which is what a biodiversity study ultimately cares about, and (2)
three cheap, classifier-free **acoustic indices** (ACI, BI, ADI) that many PAM studies use as a
biodiversity proxy. The comparison asks a methodological question that matters for any analysis
pipeline: do the cheap indices give the same spacing answer as the species community? We also examine
how the soundscape changes through the day (diel variation) and characterise the forest–pasture
microclimate with paired temperature/humidity loggers.

# Research question

> **Primary:** How does the similarity of the species community recorded by two AudioMoths change with
> the distance between them, and is that distance–community relationship different in forest than in
> pasture?
>
> **Secondary (method comparison):** Do classifier-free acoustic indices (ACI, BI, ADI) reproduce the
> same distance / habitat answer as the BirdNET species community, and how do the indices vary with
> time of day?

# Prediction

The mechanism is **distance-decay of similarity**: sound attenuates as it travels, so each recorder
samples animals within a limited *listening area*. Two nearby recorders share listening areas and
detect a similar community; as they move apart they increasingly sample different individuals,
microhabitats, and species, so the recorded community grows more **dissimilar with distance**. Habitat
structure should modulate this: forest vegetation absorbs and scatters sound, giving each recorder a
*smaller, patchier* listening area, whereas open pasture lets sound travel farther.

- **P1 — Distance-decay.** Pairwise community dissimilarity (Bray–Curtis, Jaccard) increases with
  inter-recorder distance.
- **P2 — Habitat modulates the decay.** The increase is steeper in **forest** than pasture (forest
  attenuates sound → more localised soundscapes).
- **P3 — Diel variation.** Acoustic activity varies through the day; the biological band (BI) peaks
  around dawn/day (the chorus).
- **P4 — Index concordance.** The acoustic indices reproduce the community's distance/habitat pattern.

# Study area

The study was conducted at **Tandayapa Cloud Forest Station**, Pichincha Province, Ecuador, within the
Chocó Andino Biosphere Reserve, on the western slope of the Andes (≈0.0117°, −78.6739°, ~1,680 m
a.s.l.). Two adjacent habitats were sampled:

- **Forest** — secondary montane cloud forest: dense vegetation, higher vertical structure, closed to
  semi-closed canopy, with epiphytes, ferns, bromeliads, orchids and understory. These features matter
  acoustically because dense vegetation attenuates and scatters sound.
- **Pasture** — open area dominated by non-native grasses with scattered, unidentified thorny
  shrubs/small trees and lower vegetation complexity. The open structure increases exposure to wind and
  environmental noise, which can affect sound propagation and index values.

# Experimental design

We deployed AudioMoth recorders on short linear transects **20 m inside the edge** of each habitat, at
**1.5 m height**, facing downslope (the site is on a slope), with the same protocol in both habitats and
each point marked with flagging tape. Each transect held **three recorders** at points 1–2–3. The
**spacing was changed between deployment days**, sampling three spacings — **15, 30, and 60 m** — and
each spacing was deployed on **two separate days** (replicate blocks). Recorders were **swapped between
positions** across days to reduce equipment bias.

Because each transect compares point 1↔2, 2↔3 (one spacing apart) and 1↔3 (two spacings apart), the
realised inter-recorder distances are 15, 30, 60, and 120 m. We **analyse pairs up to 60 m** and
**exclude the single 120 m pair**: it can only come from the 60 m spacing comparing points 1↔3, and
because the forest 60 m deployments lost their point-1 recorder, a 120 m pair exists **only in pasture,
on one line, with no replicate**. Dropping it keeps the forest–pasture comparison on the **same
15–60 m range** and prevents one high-leverage point from driving the result.

```{r design-table}
det %>% distinct(deploy, date, habitat, spacing_m, point_num) %>%
  count(date, spacing_m, habitat, name = "recorders") %>%
  pivot_wider(names_from = habitat, values_from = recorders, values_fill = 0) %>%
  arrange(date) %>%
  kable(caption = "Deployments: each date sampled one spacing in both habitats (recorders per habitat).")
```

# Data collection

Each AudioMoth recorded from **10:40 a.m. to 9:40 a.m. the next day**, taking **three 2-minute
recordings per hour** (2 min on / 18 min off, e.g. :00, :20, :40) — six minutes of audio per hour,
spanning the full day–night cycle (Ross et al., 2023). Data were downloaded in the field after each
session. We complemented the audio with **two Kestrel DROP 2 dataloggers** (one per habitat) logging
temperature and relative humidity every 10 minutes.

Detections were produced by **BirdNET** (range-filtered; birds and insects) and a complementary frog
classifier, each verifiable by ear and spectrogram in the project web app. Acoustic indices were
computed with **scikit-maad** (Ulloa et al., 2021). The cleaned inputs are in `data/processed/`, so
this report is fully reproducible from this document without the raw audio.

# Data analysis

All effect sizes come from **linear mixed models** (`lme4`/`lmerTest`) that carry the replication: the
**deployment day is a random intercept**, so each replicate block has its own baseline and the
distance effect is estimated *within* the replication structure rather than from pseudo-replicated
pairs. Following the course emphasis on **effect sizes over p-values**, every distance effect is
reported as its **slope (change per +10 m) with a 95% confidence interval** (plus a marginal R²); a CI
that excludes zero indicates an effect. The habitat comparison is the **forest − pasture slope
difference** with its 95% CI. Figures show the **mixed-model fixed-effect fit with 95% bands**; points
are adjusted for the deploy random intercept (so day-to-day baselines are removed) and the symbol marks
the replicate day.

**Cleaning (tidyverse).** From the BirdNET detections we drop three frog taxa that expert review judged
to be noise misclassified as frogs — listed explicitly here, not hidden in a file — and keep detections
at confidence ≥ 0.5 (with 0.25 / 0.7 as a sensitivity check). Insects were reviewed and kept.

```{r clean}
excluded_frogs <- c("Pipa pipa", "Sachatamia orejuela", "Teratohyla pulveratum")
dets <- det %>% filter(!species %in% excluded_frogs, confidence >= CONF_MIN)

# pairwise community dissimilarity (Bray–Curtis, Jaccard) for within-habitat recorder pairs, at any threshold
build_comm_pw <- function(cmin){
  d <- det %>% filter(!species %in% excluded_frogs, confidence >= cmin)
  comm_long <- d %>% count(deploy, habitat, point_num, species, name = "abund")
  samples <- comm_long %>% distinct(deploy, habitat, point_num) %>% mutate(sample = row_number())
  mat <- comm_long %>% left_join(samples, by = c("deploy","habitat","point_num")) %>%
    pivot_wider(id_cols = sample, names_from = species, values_from = abund, values_fill = 0)
  M <- as.matrix(mat[,-1]); rownames(M) <- mat$sample
  bc <- as.matrix(vegdist(M, "bray")); ja <- as.matrix(vegdist(M > 0, "jaccard"))
  pairs <- list()
  for (g in split(samples, list(samples$deploy, samples$habitat), drop = TRUE)) {
    if (nrow(g) < 2) next
    cb <- combn(seq_len(nrow(g)), 2)
    for (k in seq_len(ncol(cb))) {
      i <- g[cb[1,k],]; j <- g[cb[2,k],]
      pairs[[length(pairs)+1]] <- tibble(deploy = i$deploy, habitat = i$habitat,
        distance_m = day_spacing[i$deploy] * abs(i$point_num - j$point_num),
        bray = bc[as.character(i$sample), as.character(j$sample)],
        jacc = ja[as.character(i$sample), as.character(j$sample)])
    }
  }
  bind_rows(pairs) %>% filter(distance_m <= MAXDIST)
}

rep_of <- function(deploy) ifelse(deploy %in% c("day1","day2","day3"), "1st day", "2nd day")
mm_nakagawa_R2 <- function(m){
  varF <- var(as.numeric(predict(m, re.form = NA)))
  vc <- as.numeric(VarCorr(m)$deploy); vc <- if (length(vc) == 0) 0 else vc
  round(varF / (varF + vc + sigma(m)^2), 3)
}
# per-habitat slope per +10 m + 95% CI + marginal R², from dissim ~ distance + (1|deploy)
mm_hab <- function(d, resp){
  map_dfr(c("forest","pasture"), function(h){
    s <- filter(d, habitat == h)
    m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", resp), data = s, REML = TRUE))
    b <- fixef(m)[["distance_m"]]; se <- sqrt(vcov(m)["distance_m","distance_m"])
    lo <- (b - 1.96*se)*10; hi <- (b + 1.96*se)*10
    tibble(habitat = h, `Δ/+10m` = round(b*10,4), `CI low` = round(lo,4), `CI high` = round(hi,4),
           `R2(dist)` = mm_nakagawa_R2(m), `CI excl 0` = ifelse(lo > 0 | hi < 0, "yes","no"), n = nrow(s))
  })
}
# H2: forest − pasture slope difference from the interaction model
mm_h2 <- function(d, resp){
  m <- suppressWarnings(lmer(reformulate("distance_m * habitat + (1|deploy)", resp), data = d, REML = TRUE))
  V <- as.matrix(vcov(m)); b <- fixef(m); int <- grep(":habitat", names(b), value = TRUE)
  est <- -b[[int]]; se <- sqrt(V[int,int]); lo <- (est-1.96*se)*10; hi <- (est+1.96*se)*10
  tibble(metric = resp, `forest−pasture Δ/+10m` = round(est*10,4), `CI low` = round(lo,4),
         `CI high` = round(hi,4), `CI excl 0` = ifelse(lo > 0 | hi < 0, "yes","no"))
}
# clean distance-decay plot: model fit + 95% band + day-adjusted points, shape = replicate
mm_decay_plot <- function(d, resp, ylab){
  lns <- list(); pts <- list()
  for (h in c("forest","pasture")){
    s <- filter(d, habitat == h)
    m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", resp), data = s, REML = TRUE))
    bl <- ranef(m)$deploy[,1]; names(bl) <- rownames(ranef(m)$deploy)
    pts[[h]] <- tibble(distance_m = s$distance_m, y = s[[resp]] - bl[as.character(s$deploy)],
                       habitat = h, rep = rep_of(s$deploy))
    g <- tibble(distance_m = seq(min(s$distance_m), max(s$distance_m), length = 50))
    X <- cbind(1, g$distance_m); V <- as.matrix(vcov(m))
    g$fit <- as.numeric(X %*% fixef(m)); g$se <- sqrt(diag(X %*% V %*% t(X)))
    lns[[h]] <- mutate(g, habitat = h, lo = fit - 1.96*se, hi = fit + 1.96*se)
  }
  ggplot() +
    geom_ribbon(data = bind_rows(lns), aes(distance_m, ymin = lo, ymax = hi, fill = habitat), alpha = .15) +
    geom_line(data = bind_rows(lns), aes(distance_m, fit, color = habitat), linewidth = 1) +
    geom_point(data = bind_rows(pts), aes(distance_m, y, color = habitat, shape = rep), size = 2.6, alpha = .85) +
    scale_color_manual(values = cols) + scale_fill_manual(values = cols) +
    scale_shape_manual(values = c("1st day" = 16, "2nd day" = 17), name = "replicate") +
    labs(x = "Distance between recorders (m)", y = ylab) + theme_minimal()
}
```

# Results

## Detected richness

```{r richness}
dets %>% group_by(habitat, group) %>% summarise(n = n_distinct(species), .groups="drop") %>%
  pivot_wider(names_from = group, values_from = n, values_fill = 0) %>%
  left_join(dets %>% group_by(habitat) %>% summarise(TOTAL = n_distinct(species)), by="habitat") %>%
  kable(caption = "Detected species richness by habitat and group (confidence ≥ 0.5).")
```

Detected richness is **nearly identical** between habitats, so the contrast is about *spatial
structure*, not species count.

## Species community: distance-decay (P1, P2)

```{r community}
pw <- build_comm_pw(CONF_MIN)
mm_decay_plot(pw, "bray", "Bray–Curtis dissimilarity (day-adjusted)")
```

```{r community2, fig.cap="Species turnover (Jaccard) with distance (≤60 m). Mixed-model fit + 95% band; points day-adjusted, shape = replicate day."}
mm_decay_plot(pw, "jacc", "Jaccard dissimilarity (day-adjusted)")
```

```{r community-slopes}
slope_tab <- bind_rows(
  mm_hab(pw, "bray") %>% mutate(metric = "Bray-Curtis", .before = 1),
  mm_hab(pw, "jacc") %>% mutate(metric = "Jaccard", .before = 1))
kable(slope_tab, caption = "Per-habitat distance EFFECT SIZE (mixed model, ≤60 m): slope per +10 m, 95% CI, marginal R², n.")
```

Over the balanced 15–60 m range, community dissimilarity **increases with distance in pasture** for
both metrics, and the effect is **statistically resolved**: Bray–Curtis +0.099 per +10 m
[95% CI +0.028, +0.169] and Jaccard +0.056 [+0.004, +0.108] — **both CIs exclude zero** (P1 supported
in pasture). **Forest** shows a weak, positive-leaning trend whose CI **includes zero**. So the
distance-decay is **clear in open pasture and unresolved in forest** — the *opposite* of the
sound-attenuation prediction (P2 not supported, see below).

### Robustness to the confidence threshold

```{r sensitivity}
map_dfr(c(0.25, 0.5, 0.7), function(cm)
  mm_hab(build_comm_pw(cm), "jacc") %>% mutate(conf = cm, .before = 1) %>% select(-`R2(dist)`)) %>%
  kable(caption = "Jaccard distance slope by confidence threshold (≤60 m). The PASTURE effect excludes zero at every threshold.")
```

The **pasture** distance-decay is **robust to the threshold** (CI excludes zero at 0.25, 0.5 and 0.7).
The **forest** effect is weak and resolved only at the permissive 0.25 cutoff (which admits the most
false-positive singletons), so it is not robust.

### Does habitat change the decay rate? (P2)

```{r interaction}
bind_rows(mm_h2(pw, "bray"), mm_h2(pw, "jacc")) %>%
  kable(caption = "Habitat contrast: forest − pasture difference in distance slope (per +10 m), 95% CI. Positive = forest steeper (P2).")
```

The forest − pasture difference is **negative** (pasture steeper), the *opposite* of P2; the CI still
includes zero, so the difference is not formally resolved, but every line of evidence points to the
spatial structure being in the **open pasture**, not forest. **P2 is not supported.**

## Acoustic indices: distance-decay (P4)

```{r indices}
ip <- list()
for (g in split(idx, list(idx$deploy, idx$habitat), drop = TRUE)) {
  if (nrow(g) < 2) next
  cb <- combn(seq_len(nrow(g)), 2)
  for (k in seq_len(ncol(cb))) {
    a <- g[cb[1,k],]; b <- g[cb[2,k],]
    ip[[length(ip)+1]] <- tibble(deploy = a$deploy, habitat = a$habitat,
      distance_m = a$spacing * abs(a$point - b$point),
      dACI = abs(a$ACI-b$ACI), dBI = abs(a$BI-b$BI), dADI = abs(a$ADI-b$ADI))
  }
}
ipw <- bind_rows(ip) %>% filter(distance_m <= MAXDIST)
map_dfr(c("dACI","dBI","dADI"), function(r) mm_hab(ipw, r) %>% mutate(index = r, .before = 1)) %>%
  kable(caption = "Acoustic-index distance EFFECT SIZE (mixed model, ≤60 m): slope of |index difference| per +10 m, 95% CI.")
```

```{r index-fig, fig.width=9, fig.height=3.8, fig.cap="Per-index distance-decay (≤60 m). Mixed-model fit + 95% band; points day-adjusted, shape = replicate. All three indices are flat — no CI excludes zero."}
relab <- c(dACI = "ACI", dBI = "BI", dADI = "ADI")
pts_all <- list(); ln_all <- list()
for (r in c("dACI","dBI","dADI")) for (h in c("forest","pasture")){
  s <- filter(ipw, habitat == h)
  m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", r), data = s, REML = TRUE))
  bl <- ranef(m)$deploy[,1]; names(bl) <- rownames(ranef(m)$deploy)
  pts_all[[length(pts_all)+1]] <- tibble(index = relab[[r]], habitat = h, distance_m = s$distance_m,
                                         y = s[[r]] - bl[as.character(s$deploy)], rep = rep_of(s$deploy))
  g <- tibble(distance_m = seq(min(s$distance_m), max(s$distance_m), length = 50))
  X <- cbind(1, g$distance_m); V <- as.matrix(vcov(m))
  g$fit <- as.numeric(X %*% fixef(m)); g$se <- sqrt(diag(X %*% V %*% t(X)))
  ln_all[[length(ln_all)+1]] <- tibble(index = relab[[r]], habitat = h, distance_m = g$distance_m,
                                       fit = g$fit, lo = g$fit-1.96*g$se, hi = g$fit+1.96*g$se)
}
pts_all <- bind_rows(pts_all) %>% mutate(index = factor(index, c("ACI","BI","ADI")))
ln_all  <- bind_rows(ln_all)  %>% mutate(index = factor(index, c("ACI","BI","ADI")))
ggplot() +
  geom_ribbon(data = ln_all, aes(distance_m, ymin = lo, ymax = hi, fill = habitat), alpha = .15) +
  geom_line(data = ln_all, aes(distance_m, fit, color = habitat), linewidth = .9) +
  geom_point(data = pts_all, aes(distance_m, y, color = habitat, shape = rep), size = 2, alpha = .8) +
  scale_color_manual(values = cols) + scale_fill_manual(values = cols) +
  scale_shape_manual(values = c("1st day" = 16, "2nd day" = 17), name = "replicate") +
  facet_wrap(~index, scales = "free_y") +
  labs(x = "Distance between recorders (m)", y = "|index difference| (day-adjusted)") + theme_minimal()
```

Over 15–60 m **none of the acoustic indices shows a distance-decay** — ACI, BI and ADI all have CIs
that include zero in both habitats. The cheap indices do **not** capture the spatial turnover that the
species community resolves in pasture (P4 not supported). (In an earlier version that kept the lone
120 m pasture pair, ADI appeared significant; that signal came entirely from that single high-leverage
point and vanished once it was removed — a reminder to distrust an effect that hinges on one
observation.)

## Diel variation of the indices (P3)

```{r diel}
idxd$diel <- factor(idxd$diel, levels = c("night","dawn","day","dusk"))
map_dfr(c("ACI","BI","ADI"), function(v){
  m <- suppressWarnings(lmer(reformulate("diel*habitat + (1|device)", v), data = idxd, REML = TRUE))
  b <- fixef(m); V <- as.matrix(vcov(m))
  map_dfr(c("forest","pasture"), function(h)
    map_dfr(c("dawn","day","dusk"), function(p){
      terms <- if (h=="forest") paste0("diel",p) else c(paste0("diel",p), paste0("diel",p,":habitatpasture"))
      est <- sum(b[terms]); se <- sqrt(sum(V[terms,terms]))
      tibble(index=v, habitat=h, period=p, `vs night`=round(est,3),
             `CI low`=round(est-1.96*se,3), `CI high`=round(est+1.96*se,3),
             `CI excl 0`=ifelse((est-1.96*se)>0|(est+1.96*se)<0,"yes","no"))
    }))
}) %>% kable(caption = "Diel EFFECT SIZE: each period vs night (mixed model, recorder random intercept), 95% CI.")
```

```{r diel-fig, fig.width=9, fig.height=3.4, fig.cap="Diel variation of ACI, BI and ADI by habitat (mean ± 95% CI). BI peaks at dawn/day (the chorus); ADI dips at midday; pasture reads louder than forest."}
idxd %>% pivot_longer(c(ACI,BI,ADI), names_to="index", values_to="value") %>%
  mutate(index = factor(index, c("ACI","BI","ADI"))) %>%
  group_by(index, habitat, diel) %>%
  summarise(mean=mean(value), se=sd(value)/sqrt(n()), .groups="drop") %>%
  mutate(lo=mean-1.96*se, hi=mean+1.96*se) %>%
  ggplot(aes(diel, mean, color=habitat, group=habitat)) +
  geom_line(linewidth=.9) + geom_point(size=2.6) + geom_errorbar(aes(ymin=lo, ymax=hi), width=.15) +
  scale_color_manual(values=cols) + facet_wrap(~index, scales="free_y") +
  labs(x="Time of day", y="Index value (mean ± 95% CI)") + theme_minimal()
```

The indices clearly track **time of day**: **BI** (2–8 kHz biological band) rises at **dawn and day**
(the chorus; forest dawn +2.5, day +2.7, pasture dawn +2.0, CIs exclude zero), **ADI dips at midday**
(daytime energy concentrates into dominant frequency bands, lowering evenness), and **pasture reads
louder than forest** across the day. So the same indices that miss the subtle spatial signal capture
the strong temporal one well (P3 supported) — time of day must be balanced when comparing recorders.

## Microclimate: forest vs pasture

```{r microclimate, fig.width=9, fig.height=3.6, fig.cap="Mean temperature and relative humidity by hour (Kestrel DROP 2, June 3–7). Forest buffers both: pasture spikes hotter and drier at midday."}
clim %>% pivot_longer(c(temp_c, rh_pct), names_to="var", values_to="value") %>%
  mutate(var = recode(var, temp_c="Temperature (°C)", rh_pct="Relative humidity (%)")) %>%
  group_by(var, habitat, hour) %>% summarise(mean = mean(value), .groups="drop") %>%
  ggplot(aes(hour, mean, color=habitat)) + geom_line(linewidth=.9) + geom_point(size=1.6) +
  scale_color_manual(values=cols) + facet_wrap(~var, scales="free_y") +
  labs(x="Hour of day", y=NULL) + theme_minimal()
```

```{r microclimate-tab}
clim %>% group_by(habitat) %>%
  summarise(`mean T (°C)` = round(mean(temp_c),1), `max T` = round(max(temp_c),1),
            `mean RH (%)` = round(mean(rh_pct),1),
            `% time RH ≥ 99` = round(100*mean(rh_pct >= 99)), .groups="drop") %>%
  kable(caption = "Microclimate summary (study window). Forest is cooler and far more often saturated.")
```

The **forest buffers the microclimate**: at midday the pasture warms to ~22 °C and its humidity falls
to ~81%, while the forest stays ~19.6 °C and ~97% RH; the forest is near-saturation **84%** of the time
versus **52%** in pasture. This is consistent with the habitat-structure mechanism (closed forest
canopy moderating temperature and moisture extremes) and with the diel result that the open pasture
"sounds" busier by day.

# Discussion

Over a balanced 15–60 m range, the **species community becomes measurably more different as recorders
are spaced farther apart — in the open pasture** (P1 supported there), robustly across confidence
thresholds; in forest the same trend is weak and unresolved. Practically, by ~60 m pasture recorders
already share noticeably fewer species than adjacent ones, so spacing on the order of tens of metres
begins to buy genuinely independent community samples in open habitat.

The **habitat prediction (P2: forest steeper)** was **not supported**; if anything the steeper decay is
in pasture. The mechanistic expectation (forest attenuates sound → faster decay) is not borne out. A
plausible reading is that at 15–60 m the **pasture is the more spatially heterogeneous** habitat —
patchy edges, scattered shrubs, localised clusters of callers — so neighbouring pasture recorders
sample more distinct assemblages, while the forest interior is acoustically more uniform. This is a
tentative interpretation from a small pilot, but it is consistent across metrics and thresholds, and
the microclimate data show the forest is indeed the more homogeneous, buffered habitat.

The **method comparison (P4)** is a clean negative: **none of the acoustic indices resolves the
distance-decay that the species community does**. Indices summarise *how much* and *how spread-out* the
sound is, not *which species* are calling, so a change in community composition with distance leaves
them unmoved. The lesson is direct — **a biodiversity question about recorder spacing must be answered
with species-level data; acoustic indices alone would have wrongly concluded that spacing does not
matter**. The indices are not useless, though: they cleanly track **time of day** (P3 — the dawn/day
chorus in BI, the midday ADI dip), so they remain a cheap tool for temporal questions.

# Conclusion

At Tandayapa, over 15–60 m, the **species community recorded by two AudioMoths becomes measurably more
different as the recorders are spaced farther apart — in the open pasture**, a distance-decay robust
across confidence thresholds; in forest the trend is weak and unconfirmed, and forest did **not** decay
faster than pasture (against P2). The **acoustic indices do not detect this spatial structure at all**,
so biodiversity questions about recorder spacing must be answered with **species-level data**, not
indices alone — although the indices cleanly capture the **diel** cycle. The forest **buffers
temperature and humidity** relative to the exposed pasture. The practical message: **recorder spacing
affects sample independence and should be standardised and reported, and time of day balanced when
comparing recorders.**

# Limitations

1. **Small sample and recorder failures.** Only 10–12 recorder pairs per habitat; several AudioMoths
   failed (e.g. unit A08 damaged; A04/A06/A03 missed a day). Losses also forced us to drop the 120 m
   distance (it survived in pasture only), so the design is balanced but capped at 60 m.
2. **Spatial non-independence.** The three recorders on a transect shared points across distance
   treatments, so pairs within a deployment are not fully independent; the deploy random intercept
   mitigates but does not remove this (Sugai et al., 2019; Bradfer-Lawrence et al., 2023).
3. **Detection and index caveats.** Counts index vocal *activity*, not abundance; BirdNET can miss
   local or endemic taxa, so results should be validated by ear (Fairbairn et al., 2025; Schiavo et
   al., 2025). Indices were averaged over a sample of clips per recorder.
4. **Microclimate replication.** Only two dataloggers (one per habitat), so microclimate is described
   at the habitat level and is confounded with habitat; rain could not be separated from the forest's
   near-constant saturation.
5. **Topography.** The transects were on a slope, which can influence both sound propagation and
   recorder placement.

# Future directions

Future work should **increase the number of sampling days** and **replicate the same spacings (15, 30,
60 m) on independent transects** within each habitat (rather than extending the maximum distance), to
confirm the pasture distance-decay and reduce spatial dependence. BirdNET could be **fine-tuned with
local recordings** to improve detection of Chocó-Andino taxa (Schiavo et al., 2025). Adding more
habitat types, detailed **vegetation-structure** measurements (canopy openness, density, vertical
layers, e.g. LiDAR), and more **environmental variables** (a rain gauge, wind, leaf-wetness) would
clarify how vegetation and weather shape both sound propagation and acoustic activity (Hao et al.,
2024; Guagliumi et al., 2025).

# References

Bradfer-Lawrence, T., Gardner, N., Bunnefeld, L., Bunnefeld, N., Willis, S. G., & Dent, D. H. (2023).
Using acoustic indices in ecology: guidance on study design, analyses and interpretation. *Methods in
Ecology and Evolution.*

Fairbairn, A. J., et al. (2025). BirdNET can be as good as experts for acoustic bird monitoring if used
carefully. *PLOS ONE,* 20(10), e0330836.

Guagliumi, G., et al. (2025). The interaction between vegetation structure and acoustic environment.
*Sustainability,* 17(9), 4204.

Hao, Z., et al. (2024). Can urban forests provide acoustic refuges for birds? *Journal of Forestry
Research.*

Kahl, S., Wood, C. M., Eibl, M., & Klinck, H. (2021). BirdNET: a deep learning solution for avian
diversity monitoring. *Ecological Informatics,* 61, 101236.

Ross, S. R. P.-J., et al. (2023). Passive acoustic monitoring provides a fresh perspective on
fundamental ecological questions. *Functional Ecology,* 37(4), 959–975.

Schiavo, G., et al. (2025). Fine-tuning BirdNET for the automatic ecoacoustic monitoring of bird
species in the Italian Alpine forests. *Information,* 16(8), 628.

Sugai, L. S. M., Silva, T. S. F., Ribeiro, J. W., Jr., & Llusia, D. (2019). A roadmap for survey
designs in terrestrial acoustic monitoring. *Remote Sensing in Ecology and Conservation.*

Ulloa, J. S., Haupert, S., Latorre, J. F., Aubin, T., & Sueur, J. (2021). scikit-maad: an open-source
modular toolbox for quantitative soundscape analysis in Python. *Methods in Ecology and Evolution,*
12(12), 2334–2340.

---

```{r session, echo=FALSE}
# sessionInfo()  # uncomment for the full reproducibility footprint
```
