---
title: "When do two recorders hear the same community, and does it differ between forest and pasture?"
subtitle: "Recorder spacing and soundscape distance-decay · Tandayapa Cloud Forest Station, Ecuador · AudioMoth passive acoustic monitoring"
author: "Franz Chandi, Maria Andrade, Priscila Cajas, Noemi Castro"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    fig_width: 6.6
    fig_height: 4.4
  word_document:
    reference_docx: reference.docx
    fig_width: 6.2
    fig_height: 3.9
---

```{r setup, include=FALSE}
# Show the code in the HTML (reproducible report); hide it in the Word version (clean read).
knitr::opts_chunk$set(echo = knitr::is_html_output(), message = FALSE, warning = FALSE)
suppressMessages({
  library(tidyverse)   # dplyr, tidyr, readr, purrr, stringr, ggplot2, cleaning + plots
  library(vegan)       # community dissimilarities (Bray–Curtis, Jaccard)
  library(lme4); library(lmerTest)   # linear mixed models (deploy as random effect)
  library(knitr)
})

# Publication-style tables: plain kable in HTML, flextable (journal look) in Word.
pub_tab <- function(df, cap){
  if (knitr::is_html_output()) return(knitr::kable(df, caption = cap))
  nc <- ncol(df)
  flextable::flextable(df) |>
    flextable::set_caption(cap) |>
    flextable::border_remove() |>
    flextable::hline_top(border = officer::fp_border(color = "black", width = 2), part = "header") |>
    flextable::hline_bottom(border = officer::fp_border(color = "black", width = 1), part = "header") |>
    flextable::hline_bottom(border = officer::fp_border(color = "black", width = 2), part = "body") |>
    flextable::bold(part = "header") |>
    flextable::font(fontname = "Arial", part = "all") |>
    flextable::fontsize(size = 9, part = "all") |>
    flextable::align(j = 2:nc, align = "center", part = "all") |>
    flextable::align(j = 1, align = "left", part = "all") |>
    flextable::valign(valign = "center", part = "all") |>
    flextable::padding(padding = 4, part = "all") |>
    flextable::set_table_properties(layout = "autofit", width = 1)
}
CONF_MIN <- 0.5            # primary BirdNET confidence threshold (sensitivity to 0.25 / 0.7 below)
MAXDIST  <- 60            # analyse pairs up to 60 m (drop the lone, unreplicated 120 m pair)
cols <- c(forest = "#2C5F2D", pasture = "#D9A441")

# ---- All analyses read the COMMITTED processed data, so this report knits without the raw audio ----
PROC <- "../data/processed"
det  <- read_csv(file.path(PROC, "birdnet_detections.csv"), show_col_types = FALSE)
idx  <- read_csv(file.path(PROC, "acoustic_indices_per_recorder.csv"), show_col_types = FALSE)
idxd <- read_csv(file.path(PROC, "acoustic_indices_by_diel.csv"), show_col_types = FALSE)
clim <- read_csv(file.path(PROC, "datalogger_microclimate.csv"), show_col_types = FALSE)
day_spacing <- det %>% distinct(deploy, spacing_m) %>% deframe()
```

# Introduction

Passive acoustic monitoring (PAM) with low-cost recorders such as AudioMoth is now a standard tool
for surveying soniferous biodiversity, birds, frogs, and insects, over long periods with minimal
disturbance (Ross et al., 2023). A practical question every PAM study must answer, but rarely tests
directly, is **how far apart recorders should be placed**. If two recorders sit close together they
"hear" almost the same animals and act as one sample (pseudoreplication); if they are far apart they
capture independent slices of the community and provide genuine spatial replication. The right
spacing depends on how fast the recorded soundscape changes with distance, which should depend on the
habitat, because vegetation and terrain control how far sound travels (Bradfer-Lawrence et al., 2023).

Tandayapa spans a mosaic of dense forest and open pasture, an ideal contrast for asking whether
recorder spacing behaves differently in a structurally complex habitat versus an open one. We answer
the question in two currencies and compare them: (1) the **species community** detected by an
automated classifier (BirdNET), which is what a biodiversity study ultimately cares about, and (2)
three cheap, classifier-free **acoustic indices** (ACI, BI, ADI) that many PAM studies use as a
biodiversity proxy. The comparison asks a methodological question that matters for any analysis
pipeline: do the cheap indices give the same spacing answer as the species community? We also examine
how the soundscape changes through the day (diel variation) and characterise the forest–pasture
microclimate with paired temperature/humidity loggers.

# Research question

> **Primary:** How does the similarity of the species community recorded by two AudioMoths change with
> the distance between them, and is that distance–community relationship different in forest than in
> pasture?
>
> **Secondary (method comparison):** Do classifier-free acoustic indices (ACI, BI, ADI) reproduce the
> same distance / habitat answer as the BirdNET species community, and how do the indices vary with
> time of day?

# Prediction

The mechanism is **distance-decay of similarity**: sound attenuates as it travels, so each recorder
samples animals within a limited *listening area*. Two nearby recorders share listening areas and
detect a similar community; as they move apart they increasingly sample different individuals,
microhabitats, and species, so the recorded community grows more **dissimilar with distance**. Habitat
structure should modulate this: forest vegetation absorbs and scatters sound, giving each recorder a
*smaller, patchier* listening area, whereas open pasture lets sound travel farther.

- **P1, Distance-decay.** Pairwise community dissimilarity (Bray–Curtis, Jaccard) increases with
  inter-recorder distance.
- **P2, Habitat modulates the decay.** The increase is steeper in **forest** than pasture (forest
  attenuates sound, creating more localised soundscapes).
- **P3, Diel variation.** Acoustic activity varies through the day; the biological band (BI) peaks
  around dawn/day (the chorus).
- **P4, Index concordance.** The acoustic indices reproduce the community's distance/habitat pattern.

# Study area

The study was conducted at **Tandayapa Cloud Forest Station**, Pichincha Province, Ecuador, within the
Chocó Andino Biosphere Reserve, on the western slope of the Andes (≈0.0117°, −78.6739°, ~1,680 m
a.s.l.). Two adjacent habitats were sampled:

- **Forest**, secondary montane cloud forest: dense vegetation, higher vertical structure, closed to
  semi-closed canopy, with epiphytes, ferns, bromeliads, orchids and understory. These features matter
  acoustically because dense vegetation attenuates and scatters sound.
- **Pasture**, open area dominated by non-native grasses with scattered, unidentified thorny
  shrubs/small trees and lower vegetation complexity. The open structure increases exposure to wind and
  environmental noise, which can affect sound propagation and index values.

# Experimental design

We deployed AudioMoth recorders on short linear transects **20 m inside the edge** of each habitat, at
**1.5 m height**, facing downslope, with the same protocol in both habitats and
each point marked with flagging tape. Each transect held **three recorders** at points 1–2–3. The
**spacing was changed between deployment days**, sampling three spacings, **15, 30, and 60 m**, and
each spacing was deployed on **two separate days** (replicate blocks). Recorders were **swapped between
positions** across days to reduce equipment bias.

Because each transect compares point 1–2, 2–3 (one spacing apart) and 1–3 (two spacings apart), the
realised inter-recorder distances are 15, 30, 60, and 120 m. We **analyse pairs up to 60 m** and
**exclude the single 120 m pair**: it can only come from the 60 m spacing comparing points 1–3, and
because the forest 60 m deployments lost their point-1 recorder, a 120 m pair exists **only in pasture,
on one line, with no replicate**. Dropping it keeps the forest–pasture comparison on the **same
15–60 m range** and prevents one unusual, isolated point from driving the result.

```{r design-table}
det %>% distinct(deploy, date, habitat, spacing_m, point_num) %>%
  count(date, spacing_m, habitat, name = "recorders") %>%
  pivot_wider(names_from = habitat, values_from = recorders, values_fill = 0) %>%
  arrange(date) %>%
  pub_tab("Table 1. Deployments and working recorders per habitat.")
```

**Table 1** lists the deployments: each day used one spacing (15, 30, or 60 m) in both habitats, and
each spacing was repeated on two days.

# Data collection

Each AudioMoth recorded from **10:40 a.m. to 9:40 a.m. the next day**, taking **three 2-minute
recordings per hour** (2 min on / 18 min off, e.g. :00, :20, :40), six minutes of audio per hour,
spanning the full day–night cycle (Ross et al., 2023). Data were downloaded in the field after each
session. We complemented the audio with **two Kestrel DROP 2 dataloggers** (one per habitat) logging
temperature and relative humidity every 10 minutes.

Detections were produced by **BirdNET** (range-filtered; birds and insects) and a complementary frog
classifier, each verifiable by ear and spectrogram in the project web app. Acoustic indices were
computed with **scikit-maad** (Ulloa et al., 2021). The cleaned inputs are in `data/processed/`, so
this report is fully reproducible from this document without the raw audio.

> **Reproducibility, all code and processed data:** <https://github.com/Fr4nzz/tandayapa-soundscape>
> (clone, open the RStudio project, and knit `reports/…​.Rmd`; see the repo README).

# Data analysis

## Metrics and terms used

- **Dissimilarity**, a number from 0 to 1 that says how *different* two recorders' communities are.
  0 = identical, 1 = nothing in common. We measure it two ways:
  - **Jaccard dissimilarity**, based only on *which species* are present (presence/absence). It asks:
    do the two recorders hear the **same list of species**? It treats a species heard once the same as
    one heard 500 times, so it is sensitive to rare species and false positives.
  - **Bray–Curtis dissimilarity**, based on *how much* each species is heard (here, the number of
    detections = **vocal activity**). It asks: do the two recorders hear the same species **in similar
    amounts**? It gives less weight to rare species, so it is steadier against noise.
  - We report **both** because if they agree, the result is more trustworthy (it is not an artefact of
    one method's assumptions).
- **Distance-decay**, the idea that two recorders record **more different** communities as the
  distance between them grows.
- **Effect size**, the *size* of a relationship (e.g. how much dissimilarity rises per +10 m), not
  just whether it is "significant". We report it as a **slope**.
- **95% confidence interval (CI)**, the range of plausible values for that slope. If the CI **does not
  include zero**, the effect is statistically clear; if it **includes zero**, we cannot rule out "no
  effect".
- **R²**, the fraction of the variation the relationship explains (0 = none, 1 = all).
- **Mixed model / random effect**, a model that accounts for our sampling design: because each
  deployment **day** has its own baseline (and the three recorders of a day share points), we let each
  day have its own intercept (a "random effect") so the distance effect is estimated *fairly*, not from
  pseudo-replicated pairs.
- **ACI, BI, ADI**, three "acoustic indices" that summarise a whole recording with one number each
  (complexity, energy in the 2–8 kHz biological band, and how evenly energy is spread across
  frequencies) **without identifying species**.

## Statistical analysis

All **effect sizes** come from **linear mixed models** (`lme4`/`lmerTest`) with the **deployment day as
a random effect**. For a pair of recorders $i,j$ in habitat $h$ and deployment day $d$, the
dissimilarity is modelled as a function of the inter-recorder distance:

$$y_{ijd} = \beta_0 + \beta_1\,\text{distance}_{ij} + u_d + \varepsilon_{ijd}, \qquad u_d \sim N(0,\sigma_u^2)$$

where $\beta_1$ is the **effect size** (change in dissimilarity per metre, reported per +10 m), $u_d$ is
the random intercept for each deployment day, and the habitat comparison is the difference in $\beta_1$
between forest and pasture. The two community dissimilarities are

$$\text{Bray–Curtis}_{jk} = \frac{\sum_i |x_{ij}-x_{ik}|}{\sum_i (x_{ij}+x_{ik})}, \qquad
\text{Jaccard}_{jk} = 1 - \frac{a}{a+b+c}$$

where $x_{ij}$ is the activity (detection count) of species $i$ at recorder $j$, and for Jaccard $a$ is
the number of species shared by both recorders and $b,c$ the numbers unique to each.

Following the course emphasis on **effect sizes over p-values**, every distance effect is reported as
its **slope (per +10 m) with a 95% confidence interval** (plus an R²); a CI that excludes zero
indicates an effect. Figures show the model's fitted line with a 95% band; the points are adjusted for
the day random effect (so day-to-day baseline differences are removed and the points sit close to the
line), and the symbol marks whether the point is from the first or second day that spacing was used.

**Cleaning (tidyverse).** From the BirdNET detections we removed three frog taxa (*Pipa pipa*,
*Sachatamia orejuela*, *Teratohyla pulveratum*) that we judged to be misclassified noise after
listening to the recordings and inspecting their spectrograms, and kept detections at confidence
≥ 0.5 (with 0.25 and 0.7 as a check). Insects were reviewed and kept. A note on "abundance": our
counts are **vocal activity** (how often a species was detected), **not true abundance**: one
individual singing a lot in the forest looks like "more" than a quiet individual in the pasture, so
Bray–Curtis here is *activity-weighted*, not a real population count.

```{r clean}
excluded_frogs <- c("Pipa pipa", "Sachatamia orejuela", "Teratohyla pulveratum")
dets <- det %>% filter(!species %in% excluded_frogs, confidence >= CONF_MIN)

# pairwise community dissimilarity (Bray–Curtis, Jaccard) for within-habitat recorder pairs, at any threshold
build_comm_pw <- function(cmin){
  d <- det %>% filter(!species %in% excluded_frogs, confidence >= cmin)
  comm_long <- d %>% count(deploy, habitat, point_num, species, name = "abund")
  samples <- comm_long %>% distinct(deploy, habitat, point_num) %>% mutate(sample = row_number())
  mat <- comm_long %>% left_join(samples, by = c("deploy","habitat","point_num")) %>%
    pivot_wider(id_cols = sample, names_from = species, values_from = abund, values_fill = 0)
  M <- as.matrix(mat[,-1]); rownames(M) <- mat$sample
  bc <- as.matrix(vegdist(M, "bray")); ja <- as.matrix(vegdist(M > 0, "jaccard"))
  pairs <- list()
  for (g in split(samples, list(samples$deploy, samples$habitat), drop = TRUE)) {
    if (nrow(g) < 2) next
    cb <- combn(seq_len(nrow(g)), 2)
    for (k in seq_len(ncol(cb))) {
      i <- g[cb[1,k],]; j <- g[cb[2,k],]
      pairs[[length(pairs)+1]] <- tibble(deploy = i$deploy, habitat = i$habitat,
        distance_m = day_spacing[i$deploy] * abs(i$point_num - j$point_num),
        bray = bc[as.character(i$sample), as.character(j$sample)],
        jacc = ja[as.character(i$sample), as.character(j$sample)])
    }
  }
  bind_rows(pairs) %>% filter(distance_m <= MAXDIST)
}

rep_of <- function(deploy) ifelse(deploy %in% c("day1","day2","day3"), "1st day", "2nd day")
mm_nakagawa_R2 <- function(m){
  varF <- var(as.numeric(predict(m, re.form = NA)))
  vc <- as.numeric(VarCorr(m)$deploy); vc <- if (length(vc) == 0) 0 else vc
  round(varF / (varF + vc + sigma(m)^2), 3)
}
# per-habitat slope per +10 m + 95% CI + marginal R², from dissim ~ distance + (1|deploy)
mm_hab <- function(d, resp){
  map_dfr(c("forest","pasture"), function(h){
    s <- filter(d, habitat == h)
    m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", resp), data = s, REML = TRUE))
    b <- fixef(m)[["distance_m"]]; se <- sqrt(vcov(m)["distance_m","distance_m"])
    lo <- (b - 1.96*se)*10; hi <- (b + 1.96*se)*10
    tibble(habitat = h, `Δ/+10m` = round(b*10,4), `CI low` = round(lo,4), `CI high` = round(hi,4),
           `R2(dist)` = mm_nakagawa_R2(m), `CI excl 0` = ifelse(lo > 0 | hi < 0, "yes","no"), n = nrow(s))
  })
}
# H2: forest − pasture slope difference from the interaction model
mm_h2 <- function(d, resp){
  m <- suppressWarnings(lmer(reformulate("distance_m * habitat + (1|deploy)", resp), data = d, REML = TRUE))
  V <- as.matrix(vcov(m)); b <- fixef(m); int <- grep(":habitat", names(b), value = TRUE)
  est <- -b[[int]]; se <- sqrt(V[int,int]); lo <- (est-1.96*se)*10; hi <- (est+1.96*se)*10
  tibble(metric = resp, `forest−pasture Δ/+10m` = round(est*10,4), `CI low` = round(lo,4),
         `CI high` = round(hi,4), `CI excl 0` = ifelse(lo > 0 | hi < 0, "yes","no"))
}
# clean distance-decay plot: model fit + 95% band + day-adjusted points, shape = replicate
mm_decay_plot <- function(d, resp, ylab){
  lns <- list(); pts <- list()
  for (h in c("forest","pasture")){
    s <- filter(d, habitat == h)
    m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", resp), data = s, REML = TRUE))
    bl <- ranef(m)$deploy[,1]; names(bl) <- rownames(ranef(m)$deploy)
    pts[[h]] <- tibble(distance_m = s$distance_m, y = s[[resp]] - bl[as.character(s$deploy)],
                       habitat = h, rep = rep_of(s$deploy))
    g <- tibble(distance_m = seq(min(s$distance_m), max(s$distance_m), length = 50))
    X <- cbind(1, g$distance_m); V <- as.matrix(vcov(m))
    g$fit <- as.numeric(X %*% fixef(m)); g$se <- sqrt(diag(X %*% V %*% t(X)))
    lns[[h]] <- mutate(g, habitat = h, lo = fit - 1.96*se, hi = fit + 1.96*se)
  }
  ggplot() +
    geom_ribbon(data = bind_rows(lns), aes(distance_m, ymin = lo, ymax = hi, fill = habitat), alpha = .15) +
    geom_line(data = bind_rows(lns), aes(distance_m, fit, color = habitat), linewidth = 1) +
    geom_point(data = bind_rows(pts), aes(distance_m, y, color = habitat, shape = rep), size = 2.6, alpha = .85) +
    scale_color_manual(values = cols) + scale_fill_manual(values = cols) +
    scale_shape_manual(values = c("1st day" = 16, "2nd day" = 17), name = "replicate") +
    labs(x = "Distance between recorders (m)", y = ylab) + theme_minimal()
}
```

# Results

## Detected richness and vocal activity

```{r richness}
dets %>% group_by(habitat, group) %>% summarise(n = n_distinct(species), .groups="drop") %>%
  pivot_wider(names_from = group, values_from = n, values_fill = 0) %>%
  left_join(dets %>% group_by(habitat) %>% summarise(TOTAL = n_distinct(species)), by="habitat") %>%
  pub_tab("Table 2. Species detected by habitat and group.")
```

**Table 2** shows how many species we detected. Detected richness is **nearly identical** between
forest and pasture, so the spacing contrast is about *spatial structure*, not species count.

**Figure 1** puts richness (how *many* species) next to vocal activity (how *often* they were
detected). The two habitats hold a similar number of species, but they differ in how much each group
calls. Remember (see Data analysis) that vocal activity is **not true abundance**: the same individual
may simply sing more in one habitat than the other, so we read the right-hand panel as "how much sound
each group produced", not "how many animals there were". Pasture recorders logged **more detections**
than forest ones, especially for frogs and birds. This does not mean the pasture is richer (richness is
similar, and forest holds slightly more bird species): it more likely reflects more calling in the open
and sound travelling farther where there is less vegetation to absorb it. It is therefore not a reason
to place recorders only in open areas; richness and community composition, not detection counts, are
what a biodiversity survey should compare.

```{r richness-activity-fig, fig.width=9, fig.height=3.8, fig.cap="Figure 1. (A) Species richness and (B) vocal activity (detections) by habitat and group."}
rich <- dets %>% group_by(habitat, group) %>% summarise(value = n_distinct(species), .groups="drop") %>% mutate(panel = "A. Species richness")
act  <- dets %>% count(habitat, group, name = "value") %>% mutate(panel = "B. Vocal activity (detections)")
bind_rows(rich, act) %>%
  ggplot(aes(group, value, fill = habitat)) +
  geom_col(position = position_dodge(width = .8), width = .7) +
  scale_fill_manual(values = cols) + facet_wrap(~panel, scales = "free_y") +
  labs(x = NULL, y = NULL) + theme_minimal()
```

## Species community: distance-decay (P1, P2)

In **Figures 2 and 3**, an upward line means the two recorders record **more different** communities as
they get farther apart (distance-decay). The gold line is pasture, the green line is forest; the shaded
band is the 95% CI, and the circle/triangle symbols mark the first/second replicate day.

```{r community, fig.cap="Figure 2. Bray–Curtis community dissimilarity vs distance, forest vs pasture (≤60 m)."}
pw <- build_comm_pw(CONF_MIN)
mm_decay_plot(pw, "bray", "Bray–Curtis dissimilarity (day-adjusted)")
```

```{r community2, fig.cap="Figure 3. Jaccard species turnover vs distance (≤60 m)."}
mm_decay_plot(pw, "jacc", "Jaccard dissimilarity (day-adjusted)")
```

```{r community-slopes}
slope_tab <- bind_rows(
  mm_hab(pw, "bray") %>% mutate(metric = "Bray-Curtis", .before = 1),
  mm_hab(pw, "jacc") %>% mutate(metric = "Jaccard", .before = 1))
pub_tab(slope_tab, "Table 3. Distance effect on community dissimilarity by habitat (mixed model, ≤60 m).")
```

In Table 3, the column **Δ/+10m** is the effect size: how much more different two recorders become for
every extra 10 m apart, where a positive number means distance-decay. The **CI low / CI high** columns
give the plausible range, and **CI excl 0 = yes** means the effect is statistically clear (the range
does not include zero).

The result: over 15–60 m, community difference **increases with distance in the pasture** for both
measures, and the effect is **clear**, Bray–Curtis +0.099 per +10 m (CI +0.028 to +0.169) and Jaccard
+0.056 (CI +0.004 to +0.108), both with "CI excl 0 = yes" (Figures 2–3, Table 3). In the **forest** the
trend is weak and its CI includes zero. So the distance-decay is **clear in the open pasture and
unresolved in forest**, the *opposite* of our sound-attenuation prediction (P2; see below).

### Robustness to the confidence threshold

```{r sensitivity}
map_dfr(c(0.25, 0.5, 0.7), function(cm)
  mm_hab(build_comm_pw(cm), "jacc") %>% mutate(conf = cm, .before = 1) %>% select(-`R2(dist)`)) %>%
  pub_tab("Table 4. Jaccard distance slope at three confidence thresholds (≤60 m).")
```

**Table 4** checks whether the result depends on how strict we are with BirdNET's confidence. The
**pasture** distance-decay holds at **every** threshold (0.25, 0.5, 0.7, "CI excl 0 = yes" each time),
so it is robust. The **forest** effect appears only at the loosest cut-off (0.25, which lets in the most
false positives), so it is not reliable.

### Does habitat change the decay rate? (P2)

```{r interaction}
bind_rows(mm_h2(pw, "bray"), mm_h2(pw, "jacc")) %>%
  pub_tab("Table 5. Forest − pasture difference in distance slope (per +10 m), with 95% CI.")
```

**Table 5** directly tests P2 (is forest steeper than pasture?). The forest − pasture difference is
**negative**, meaning **pasture** is the steeper habitat, the *opposite* of what we predicted. The CI
still includes zero (so it is not formally proven), but together with Table 3 everything points to the
spatial signal being in the **open pasture**, not the forest. **P2 is not supported.**

## Acoustic indices: distance-decay (P4)

```{r indices}
ip <- list()
for (g in split(idx, list(idx$deploy, idx$habitat), drop = TRUE)) {
  if (nrow(g) < 2) next
  cb <- combn(seq_len(nrow(g)), 2)
  for (k in seq_len(ncol(cb))) {
    a <- g[cb[1,k],]; b <- g[cb[2,k],]
    ip[[length(ip)+1]] <- tibble(deploy = a$deploy, habitat = a$habitat,
      distance_m = a$spacing * abs(a$point - b$point),
      dACI = abs(a$ACI-b$ACI), dBI = abs(a$BI-b$BI), dADI = abs(a$ADI-b$ADI))
  }
}
ipw <- bind_rows(ip) %>% filter(distance_m <= MAXDIST)
map_dfr(c("dACI","dBI","dADI"), function(r) mm_hab(ipw, r) %>% mutate(index = r, .before = 1)) %>%
  pub_tab("Table 6. Distance effect on each acoustic index difference (mixed model, ≤60 m).")
```

Table 6 applies the same test to the indices: for each one (ACI, BI, ADI) we measured how different two
recorders' values are and whether that difference grows with distance. If an index captured the spatial
pattern, its **CI excl 0** would read **yes**; here every row reads **no** (all CIs include zero), so
the indices do not change with distance. They are flat (Figure 4) and miss the spatial signal that the
species community shows.

```{r index-fig, fig.width=9, fig.height=3.8, fig.cap="Figure 4. Difference in each acoustic index vs distance, forest vs pasture (≤60 m)."}
relab <- c(dACI = "ACI", dBI = "BI", dADI = "ADI")
pts_all <- list(); ln_all <- list()
for (r in c("dACI","dBI","dADI")) for (h in c("forest","pasture")){
  s <- filter(ipw, habitat == h)
  m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", r), data = s, REML = TRUE))
  bl <- ranef(m)$deploy[,1]; names(bl) <- rownames(ranef(m)$deploy)
  pts_all[[length(pts_all)+1]] <- tibble(index = relab[[r]], habitat = h, distance_m = s$distance_m,
                                         y = s[[r]] - bl[as.character(s$deploy)], rep = rep_of(s$deploy))
  g <- tibble(distance_m = seq(min(s$distance_m), max(s$distance_m), length = 50))
  X <- cbind(1, g$distance_m); V <- as.matrix(vcov(m))
  g$fit <- as.numeric(X %*% fixef(m)); g$se <- sqrt(diag(X %*% V %*% t(X)))
  ln_all[[length(ln_all)+1]] <- tibble(index = relab[[r]], habitat = h, distance_m = g$distance_m,
                                       fit = g$fit, lo = g$fit-1.96*g$se, hi = g$fit+1.96*g$se)
}
pts_all <- bind_rows(pts_all) %>% mutate(index = factor(index, c("ACI","BI","ADI")))
ln_all  <- bind_rows(ln_all)  %>% mutate(index = factor(index, c("ACI","BI","ADI")))
ggplot() +
  geom_ribbon(data = ln_all, aes(distance_m, ymin = lo, ymax = hi, fill = habitat), alpha = .15) +
  geom_line(data = ln_all, aes(distance_m, fit, color = habitat), linewidth = .9) +
  geom_point(data = pts_all, aes(distance_m, y, color = habitat, shape = rep), size = 2, alpha = .8) +
  scale_color_manual(values = cols) + scale_fill_manual(values = cols) +
  scale_shape_manual(values = c("1st day" = 16, "2nd day" = 17), name = "replicate") +
  facet_wrap(~index, scales = "free_y") +
  labs(x = "Distance between recorders (m)", y = "|index difference| (day-adjusted)") + theme_minimal()
```

Over 15–60 m **none of the acoustic indices shows a distance-decay**, ACI, BI and ADI all have CIs
that include zero in both habitats. The cheap indices do **not** capture the spatial turnover that the
species community resolves in pasture (P4 not supported). Including the single 120 m pasture pair would
produce an apparent ADI signal, but that signal depends entirely on that one unreplicated point, which
is why it was excluded (Methods).

## Diel variation of the indices (P3)

Here we keep it simple and show the pattern in one figure (the diel effect is large and obvious, so we
do not need a separate effect-size table as for the main distance question).

```{r diel-fig, fig.width=9, fig.height=3.4, fig.cap="Figure 5. Diel variation of ACI, BI and ADI by habitat (mean ± 95% CI)."}
idxd$diel <- factor(idxd$diel, levels = c("night","dawn","day","dusk"))
idxd %>% pivot_longer(c(ACI,BI,ADI), names_to="index", values_to="value") %>%
  mutate(index = factor(index, c("ACI","BI","ADI"))) %>%
  group_by(index, habitat, diel) %>%
  summarise(mean=mean(value), se=sd(value)/sqrt(n()), .groups="drop") %>%
  mutate(lo=mean-1.96*se, hi=mean+1.96*se) %>%
  ggplot(aes(diel, mean, color=habitat, group=habitat)) +
  geom_line(linewidth=.9) + geom_point(size=2.6) + geom_errorbar(aes(ymin=lo, ymax=hi), width=.15) +
  scale_color_manual(values=cols) + facet_wrap(~index, scales="free_y") +
  labs(x="Time of day", y="Index value (mean ± 95% CI)") + theme_minimal()
```

The indices clearly track **time of day** (Figure 5): **BI** (the biological 2–8 kHz band) rises at
**dawn and day**, the dawn chorus, and falls at night; **ADI dips at midday** (by day the sound
concentrates into a few dominant frequencies, so it is less "even"); and **pasture sounds louder than
forest** through the day. So the same indices that miss the subtle *spatial* signal capture the strong
*time-of-day* signal well (P3 supported), which is why time of day must be balanced when comparing
recorders.

## Microclimate: forest vs pasture

```{r microclimate, fig.width=9, fig.height=3.6, fig.cap="Figure 6. Mean temperature and humidity by hour, forest vs pasture (June 3–7)."}
clim %>% pivot_longer(c(temp_c, rh_pct), names_to="var", values_to="value") %>%
  mutate(var = recode(var, temp_c="Temperature (°C)", rh_pct="Relative humidity (%)")) %>%
  group_by(var, habitat, hour) %>% summarise(mean = mean(value), .groups="drop") %>%
  ggplot(aes(hour, mean, color=habitat)) + geom_line(linewidth=.9) + geom_point(size=1.6) +
  scale_color_manual(values=cols) + facet_wrap(~var, scales="free_y") +
  labs(x="Hour of day", y=NULL) + theme_minimal()
```

```{r microclimate-tab}
clim %>% group_by(habitat) %>%
  summarise(`mean T (°C)` = round(mean(temp_c),1), `max T` = round(max(temp_c),1),
            `mean RH (%)` = round(mean(rh_pct),1),
            `% time RH ≥ 99` = round(100*mean(rh_pct >= 99)), .groups="drop") %>%
  pub_tab("Table 7. Microclimate summary by habitat (study window).")
```

The **forest buffers the microclimate** (Figure 6, Table 7): at midday the pasture warms to ~22 °C and
its humidity falls to ~81%, while the forest stays ~19.6 °C and ~97% humidity; the forest is near
100% humidity **84%** of the time versus **52%** in the pasture. This fits the idea that the closed
forest canopy softens temperature and moisture extremes, and it matches the diel result that the open
pasture "sounds" busier by day.

# Discussion

Over a balanced 15–60 m range, the **species community becomes measurably more different as recorders
are spaced farther apart, in the open pasture** (P1 supported there), robustly across confidence
thresholds; in forest the same trend is weak and unresolved. Practically, by ~60 m pasture recorders
already share noticeably fewer species than adjacent ones, so spacing on the order of tens of metres
begins to buy genuinely independent community samples in open habitat.

The **habitat prediction (P2: forest steeper)** was **not supported**; if anything the steeper decay is
in pasture. The mechanistic expectation (forest attenuates sound, giving faster decay) is not borne out. A
plausible reading is that at 15–60 m the **pasture is the more spatially heterogeneous** habitat, 
patchy edges, scattered shrubs, localised clusters of callers, so neighbouring pasture recorders
sample more distinct assemblages, while the forest interior is acoustically more uniform. This is a
tentative interpretation from a small pilot, but it is consistent across metrics and thresholds, and
the microclimate data show the forest is indeed the more homogeneous, buffered habitat.

The **method comparison (P4)** is a clean negative: **none of the acoustic indices resolves the
distance-decay that the species community does**. Indices summarise *how much* and *how spread-out* the
sound is, not *which species* are calling, so a change in community composition with distance leaves
them unmoved. The lesson is direct, **a biodiversity question about recorder spacing must be answered
with species-level data; acoustic indices alone would have wrongly concluded that spacing does not
matter**. The indices are not useless, though: they cleanly track **time of day** (P3, the dawn/day
chorus in BI, the midday ADI dip), so they remain a cheap tool for temporal questions.

# Conclusion

At Tandayapa, over 15–60 m, the **species community recorded by two AudioMoths becomes measurably more
different as the recorders are spaced farther apart, in the open pasture**, a distance-decay robust
across confidence thresholds; in forest the trend is weak and unconfirmed, and forest did **not** decay
faster than pasture (against P2). The **acoustic indices do not detect this spatial structure at all**,
so biodiversity questions about recorder spacing must be answered with **species-level data**, not
indices alone, although the indices cleanly capture the **diel** cycle. The forest **buffers
temperature and humidity** relative to the exposed pasture. The practical message: **recorder spacing
affects sample independence and should be standardised and reported, and time of day balanced when
comparing recorders.**

# Limitations

1. **Small sample and recorder failures.** Only 10–12 recorder pairs per habitat; several AudioMoths
   failed (e.g. unit A08 damaged; A04/A06/A03 missed a day). Losses also forced us to drop the 120 m
   distance (it survived in pasture only), so the design is balanced but capped at 60 m.
2. **Spatial non-independence.** The three recorders on a transect shared points across distance
   treatments, so pairs within a deployment are not fully independent; the deploy random intercept
   mitigates but does not remove this (Sugai et al., 2019; Bradfer-Lawrence et al., 2023).
3. **Detection and index caveats.** Counts index vocal *activity*, not abundance; BirdNET can miss
   local or endemic taxa, so results should be validated by ear (Fairbairn et al., 2025; Schiavo et
   al., 2025). Indices were averaged over a sample of clips per recorder.
4. **Microclimate replication.** Only two dataloggers (one per habitat), so microclimate is described
   at the habitat level and is confounded with habitat; rain could not be separated from the forest's
   near-constant saturation.
5. **Topography.** The transects were on a slope, which can influence both sound propagation and
   recorder placement.

# Future directions

Future work should **increase the number of sampling days** and **replicate the same spacings (15, 30,
60 m) on independent transects** within each habitat (rather than extending the maximum distance), to
confirm the pasture distance-decay and reduce spatial dependence. BirdNET could be **fine-tuned with
local recordings** to improve detection of Chocó-Andino taxa (Schiavo et al., 2025). Adding more
habitat types, detailed **vegetation-structure** measurements (canopy openness, density, vertical
layers, e.g. LiDAR), and more **environmental variables** (a rain gauge, wind, leaf-wetness) would
clarify how vegetation and weather shape both sound propagation and acoustic activity (Hao et al.,
2024; Guagliumi et al., 2025).

# References

Bradfer-Lawrence, T., Gardner, N., Bunnefeld, L., Bunnefeld, N., Willis, S. G., & Dent, D. H. (2023).
Using acoustic indices in ecology: guidance on study design, analyses and interpretation. *Methods in
Ecology and Evolution.*

Fairbairn, A. J., et al. (2025). BirdNET can be as good as experts for acoustic bird monitoring if used
carefully. *PLOS ONE,* 20(10), e0330836.

Guagliumi, G., et al. (2025). The interaction between vegetation structure and acoustic environment.
*Sustainability,* 17(9), 4204.

Hao, Z., et al. (2024). Can urban forests provide acoustic refuges for birds? *Journal of Forestry
Research.*

Kahl, S., Wood, C. M., Eibl, M., & Klinck, H. (2021). BirdNET: a deep learning solution for avian
diversity monitoring. *Ecological Informatics,* 61, 101236.

Ross, S. R. P.-J., et al. (2023). Passive acoustic monitoring provides a fresh perspective on
fundamental ecological questions. *Functional Ecology,* 37(4), 959–975.

Schiavo, G., et al. (2025). Fine-tuning BirdNET for the automatic ecoacoustic monitoring of bird
species in the Italian Alpine forests. *Information,* 16(8), 628.

Sugai, L. S. M., Silva, T. S. F., Ribeiro, J. W., Jr., & Llusia, D. (2019). A roadmap for survey
designs in terrestrial acoustic monitoring. *Remote Sensing in Ecology and Conservation.*

Ulloa, J. S., Haupert, S., Latorre, J. F., Aubin, T., & Sueur, J. (2021). scikit-maad: an open-source
modular toolbox for quantitative soundscape analysis in Python. *Methods in Ecology and Evolution,*
12(12), 2334–2340.

---

```{r session, echo=FALSE}
# sessionInfo()  # uncomment for the full reproducibility footprint
```
