---
title: "When do two recorders hear the same community, and does it differ between forest and pasture?"
subtitle: "Recorder spacing and soundscape distance-decay · Tandayapa Cloud Forest Station, Ecuador · AudioMoth passive acoustic monitoring"
author: "Franz Chandi, Maria Andrade, Priscila Cajas, Noemi Castro"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    fig_width: 6.6
    fig_height: 4.4
  word_document:
    reference_docx: reference.docx
    fig_width: 6.2
    fig_height: 3.9
---

```{r setup, include=FALSE}
# Show the code in the HTML (reproducible report); hide it in the Word version (clean read).
knitr::opts_chunk$set(echo = knitr::is_html_output(), message = FALSE, warning = FALSE)
suppressMessages({
  library(tidyverse)   # dplyr, tidyr, readr, purrr, stringr, ggplot2: cleaning + plots
  library(vegan)       # community dissimilarities (Bray-Curtis, Jaccard)
  library(lme4); library(lmerTest)   # linear mixed models
  library(knitr)
})
CONF_MIN <- 0.5; MAXDIST <- 60
cols <- c(forest = "#2C5F2D", pasture = "#D9A441")

# committed processed data, so the report knits without the raw audio
PROC <- "../data/processed"
det  <- read_csv(file.path(PROC, "birdnet_detections.csv"), show_col_types = FALSE)
idx  <- read_csv(file.path(PROC, "acoustic_indices_per_recorder.csv"), show_col_types = FALSE)
idxd <- read_csv(file.path(PROC, "acoustic_indices_by_diel.csv"), show_col_types = FALSE)
clim <- read_csv(file.path(PROC, "datalogger_microclimate.csv"), show_col_types = FALSE)
day_spacing <- det %>% distinct(deploy, spacing_m) %>% deframe()
excluded_frogs <- c("Pipa pipa", "Sachatamia orejuela", "Teratohyla pulveratum")

# Publication-style tables: kable in HTML, flextable (journal look) in Word.
pub_tab <- function(df, cap){
  if (knitr::is_html_output()) return(knitr::kable(df, caption = cap))
  nc <- ncol(df)
  flextable::flextable(df) |> flextable::set_caption(cap) |> flextable::border_remove() |>
    flextable::hline_top(border = officer::fp_border(color="black", width=2), part="header") |>
    flextable::hline_bottom(border = officer::fp_border(color="black", width=1), part="header") |>
    flextable::hline_bottom(border = officer::fp_border(color="black", width=2), part="body") |>
    flextable::bold(part="header") |> flextable::font(fontname="Arial", part="all") |>
    flextable::fontsize(size=9, part="all") |> flextable::align(j=2:nc, align="center", part="all") |>
    flextable::align(j=1, align="left", part="all") |> flextable::padding(padding=4, part="all") |>
    flextable::set_table_properties(layout="autofit", width=1)
}
rep_of <- function(deploy) ifelse(deploy %in% c("day1","day2","day3"), "1st day", "2nd day")
```

# Introduction

Tropical montane cloud forests are among the most biodiverse ecosystems on Earth, with high richness
and endemism (Bruijnzeel et al., 2011); Ecuador's Chocó Andino is a global biodiversity hotspot, now a
mosaic of forest remnants and human-modified habitats such as pasture (UNESCO, 2024). Soundscape
ecology holds that the sounds of an environment carry information about its biological communities and
ecological processes (Pijanowski et al., 2011; Farina, 2014), and passive acoustic monitoring (PAM) is
now widely used to track biodiversity across space and time (Shonfield & Bayne, 2017; Sugai et al.,
2019). **Acoustic indices** such as the Acoustic Complexity Index (ACI), Bioacoustic Index (BI) and
Acoustic Diversity Index (ADI) summarise soundscape structure and allow comparisons among habitats and
conditions (Sueur et al., 2014; Metcalf et al., 2021).

A recorder samples only a portion of the surrounding soundscape, and how large that portion is depends
on sound propagation and detectability, which vary with call frequency, noise, weather and habitat
structure (Haupert et al., 2023; Winiarska et al., 2024). Recorders placed close together may
repeatedly detect the same sources, while recorders far apart sample different parts of the soundscape,
so the distance at which recordings begin to diverge is central to designing representative monitoring
and remains poorly known in tropical montane systems (Sugai et al., 2019). Habitat structure should
modulate this: the **acoustic adaptation hypothesis** holds that habitat physically shapes how signals
propagate (Morton, 1975), and forest vegetation absorbs and scatters sound, shortening detection
distances relative to open habitat (Entling et al., 2025).

We therefore ask: **at what distance do two AudioMoth recorders begin to capture different acoustic
information, and does this distance differ between forest and pasture?** We measure acoustic difference
primarily with the indices (ACI, BI, ADI) and complement them with the species community recovered by
an automated classifier (BirdNET), to test whether the cheap indices capture the same spatial structure
as species turnover. We hypothesise that **(H1)** acoustic difference increases with inter-recorder
distance; **(H2)** the increase is steeper (a larger slope) in forest than pasture, because dense
vegetation attenuates sound more strongly; and **(H3)** differences are largest during periods of high
biological activity (the dawn/day chorus). Accordingly we predict that $|\Delta$ACI$|$, $|\Delta$BI$|$
and $|\Delta$ADI$|$ rise with distance, more so in forest, and that between-site differences peak during
the chorus.

# Study area

The study was conducted at **Tandayapa Cloud Forest Station**, Pichincha Province, Ecuador, within the
Chocó Andino Biosphere Reserve, on the western slope of the Andes (≈0.0117°, −78.6739°, ~1,680 m
a.s.l.). Two adjacent habitats were sampled: **forest** (secondary montane cloud forest, dense
vegetation with closed to semi-closed canopy, epiphytes, ferns and understory, which attenuates and
scatters sound) and **pasture** (open, dominated by non-native grasses with scattered unidentified
shrubs, more exposed to wind and environmental noise).

![Figure 1. (A) Trail from Tandayapa Station to Domos, with the sampling area shown as small green (forest) and yellow (pasture) points. (B) Zoomed view of the sampling points with recorder distances from the forest edge. Forest points are raw GPS coordinates; pasture points are corrected to the design positions (0, 15, 30, 60, 120 m from the edge).](../outputs/analysis/sampling_map_merged.png)

# Experimental design

We deployed AudioMoth recorders on short linear transects **20 m inside the edge** of each habitat, at
**1.5 m height**, facing downslope, with the same protocol in both habitats and each point marked with
flagging tape. Each transect held **three recorders** at points 1–2–3. The **spacing was changed
between deployment days**, sampling **15, 30 and 60 m**, and each spacing was deployed on **two separate
days** (replicate blocks). Recorders were **swapped between positions** across days to reduce equipment
bias. Because each transect compares points 1–2 and 2–3 (one spacing apart) and 1–3 (two spacings
apart), the realised distances are 15, 30, 60 and 120 m; we **analyse pairs up to 60 m** and exclude the
single 120 m pair, which existed in pasture only, on one line, with no replicate (forest lost its 60 m
point-1 recorder). This keeps the forest–pasture comparison on the same 15–60 m range.

```{r design-table, include=FALSE}
# Table 1 is now presented as text below
deploy_summary <- det %>% distinct(deploy, date, habitat, spacing_m, point_num) %>%
  count(date, spacing_m, habitat, name = "recorders") %>%
  arrange(date)
```

Six deployment days were used (2–7 June 2026), with two days per spacing (15 m, 30 m, 60 m). Each day placed three recorders per habitat (forest and pasture), giving 6 recorders per day when all units worked. Recorders were swapped between positions across days to reduce equipment bias. Day 1 (2 June) used 15 m spacing with 2 forest and 3 pasture recorders; day 2 (3 June) used 30 m with 3 forest and 3 pasture; day 3 (4 June) used 60 m with 2 forest and 3 pasture; day 4 (5 June) used 15 m with 3 forest and 3 pasture; day 5 (6 June) used 30 m with 3 forest and 3 pasture; and day 6 (7 June) used 60 m with 2 forest and 3 pasture.

# Data collection

For analysis we used the window from **11:00 a.m. to 9:40 a.m. the next day**, discarding the
surrounding period when recorders were placed, collected, and swapped between positions, so that
handling was not counted as data. Within this window each recorder took **three 2-minute recordings per
hour** (2 min on / 18 min off), six minutes of audio per hour, covering the full day–night cycle (Ross
et al., 2023). We also ran **two Kestrel DROP 2 dataloggers** (one per habitat) logging temperature and
humidity every 10 minutes. Acoustic indices were computed with **scikit-maad** (Ulloa et al., 2021), and
species detections with **BirdNET** (range-filtered; birds and insects) and a frog classifier (Kahl et
al., 2021), each verifiable by ear and spectrogram. Cleaned inputs are in `data/processed/`, so this
report is fully reproducible without the raw audio. **All code and data:**
<https://github.com/Fr4nzz/tandayapa-soundscape>.

# Data analysis

We measure the acoustic difference between two recorders two ways. For the **indices**, the response is
the absolute difference in each index, $|\Delta$ACI$|$, $|\Delta$BI$|$, $|\Delta$ADI$|$. For the
**species community** (BirdNET birds, frogs, insects; three suspected false-positive frog taxa removed
after listening and checking spectrograms, kept at confidence ≥ 0.5), the response is the pairwise
**Bray–Curtis** dissimilarity (activity-weighted, where activity = detection count, not true abundance)
or **Jaccard** dissimilarity (presence/absence). All effect sizes come from **linear mixed models**
(`lme4`/`lmerTest`). We report **effect sizes (slopes with 95% confidence intervals)** rather than
p-values, because the size and direction of an effect are more informative than significance alone,
especially in a small study; a CI that excludes zero indicates an effect.

Two complementary models are used. For **distance**, we use one value per recorder pair so the
pair-level distance is tested cleanly:

$$\text{Difference} \sim \text{Distance} \times \text{Habitat} + (1 \mid \text{day})$$

**Why the difference?** Distance is a property of a *pair* of recorders, not of a single recorder. A recorder at 15 m has no distance on its own; distance only exists when you compare two recorders. Therefore, a distance-decay model must use the pairwise difference between the two recordings, just as a Bray–Curtis or Jaccard dissimilarity is also computed pairwise. The raw indices *are* used, but in the complementary model that asks how the absolute level of each index changes with time of day and habitat, not with distance.

For **time of day**, we compute the difference per pair within each period (dawn, day, dusk, night) and
add the pair as a random effect, which is what makes time of day estimable within pairs:

$$\text{Difference} \sim \text{Time of day} + (1 \mid \text{pair}) + (1 \mid \text{day})$$

```{r helpers}
mm_nakagawa_R2 <- function(m){
  vF <- var(as.numeric(predict(m, re.form = NA)))
  vc <- as.numeric(VarCorr(m)$deploy); vc <- if (length(vc)==0) 0 else vc
  round(vF/(vF + vc + sigma(m)^2), 3)
}
# per-habitat distance slope (+/-10 m) + 95% CI, from difference ~ distance + (1|deploy)
mm_hab <- function(d, resp){
  map_dfr(c("forest","pasture"), function(h){
    s <- filter(d, habitat==h)
    m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", resp), data=s, REML=TRUE))
    b <- fixef(m)[["distance_m"]]; se <- sqrt(vcov(m)["distance_m","distance_m"])
    lo <- (b-1.96*se)*10; hi <- (b+1.96*se)*10
    tibble(habitat=h, `Δ/+10m`=round(b*10,4), `CI low`=round(lo,4), `CI high`=round(hi,4),
           `R2`=mm_nakagawa_R2(m), `CI excl 0`=ifelse(lo>0|hi<0,"yes","no"), n=nrow(s))
  })
}
# forest - pasture difference in distance slope (the habitat effect, H2)
mm_h2 <- function(d, resp){
  m <- suppressWarnings(lmer(reformulate("distance_m*habitat + (1|deploy)", resp), data=d, REML=TRUE))
  V <- as.matrix(vcov(m)); b <- fixef(m); int <- grep(":habitat", names(b), value=TRUE)
  est <- -b[[int]]; se <- sqrt(V[int,int]); lo <- (est-1.96*se)*10; hi <- (est+1.96*se)*10
  tibble(metric=resp, `forest−pasture Δ/+10m`=round(est*10,4), `CI low`=round(lo,4),
         `CI high`=round(hi,4), `CI excl 0`=ifelse(lo>0|hi<0,"yes","no"))
}
# clean distance plot: model fit + 95% band + day-adjusted points, shape = replicate
mm_decay_plot <- function(d, resp, ylab){
  L <- list(); P <- list()
  for (h in c("forest","pasture")){
    s <- filter(d, habitat==h)
    m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", resp), data=s, REML=TRUE))
    bl <- ranef(m)$deploy[,1]; names(bl) <- rownames(ranef(m)$deploy)
    P[[h]] <- tibble(distance_m=s$distance_m, y=s[[resp]]-bl[as.character(s$deploy)], habitat=h, rep=rep_of(s$deploy))
    g <- tibble(distance_m=seq(min(s$distance_m), max(s$distance_m), length=50))
    X <- cbind(1, g$distance_m); V <- as.matrix(vcov(m))
    g$fit <- as.numeric(X %*% fixef(m)); g$se <- sqrt(diag(X %*% V %*% t(X)))
    L[[h]] <- mutate(g, habitat=h, lo=fit-1.96*se, hi=fit+1.96*se)
  }
  ggplot() +
    geom_ribbon(data=bind_rows(L), aes(distance_m, ymin=lo, ymax=hi, fill=habitat), alpha=.15) +
    geom_line(data=bind_rows(L), aes(distance_m, fit, color=habitat), linewidth=1) +
    geom_point(data=bind_rows(P), aes(distance_m, y, color=habitat, shape=rep), size=2.4, alpha=.85) +
    scale_color_manual(values=cols)+scale_fill_manual(values=cols)+
    scale_shape_manual(values=c("1st day"=16,"2nd day"=17), name="replicate")+
    labs(x="Distance between recorders (m)", y=ylab)+theme_minimal()
}
# time-of-day effects (each period vs night) from a model with a 'diel' term
tod_tab <- function(m){
  b <- fixef(m); V <- as.matrix(vcov(m))
  map_dfr(c("dawn","day","dusk"), function(p){
    nm <- paste0("diel",p); est <- b[[nm]]; se <- sqrt(V[nm,nm]); lo <- est-1.96*se; hi <- est+1.96*se
    tibble(period=p, `vs night`=round(est,3), `CI low`=round(lo,3), `CI high`=round(hi,3),
           `CI excl 0`=ifelse(lo>0|hi<0,"yes","no"))
  })
}
```

```{r aic-table}
# Compare additive vs interactive distance model for combined acoustic dissimilarity
ip <- list()
for (g in split(idx, list(idx$deploy, idx$habitat), drop=TRUE)) {
  if (nrow(g) < 2) next
  cb <- combn(seq_len(nrow(g)), 2)
  for (k in seq_len(ncol(cb))) { a <- g[cb[1,k],]; b <- g[cb[2,k],]
    ip[[length(ip)+1]] <- tibble(deploy=a$deploy, habitat=a$habitat,
      distance_m=a$spacing*abs(a$point-b$point),
      dACI=abs(a$ACI-b$ACI), dBI=abs(a$BI-b$BI), dADI=abs(a$ADI-b$ADI)) }
}
ipw <- bind_rows(ip) %>% filter(distance_m <= MAXDIST)
# z-score each index, then average into a combined dissimilarity
ipw <- ipw %>% mutate(zACI = scale(dACI)[,1], zBI = scale(dBI)[,1], zADI = scale(dADI)[,1],
                      combined_dissim = (zACI + zBI + zADI)/3)
mod_add <- suppressWarnings(lmer(combined_dissim ~ distance_m + habitat + (1|deploy), data=ipw, REML=FALSE))
mod_int <- suppressWarnings(lmer(combined_dissim ~ distance_m * habitat + (1|deploy), data=ipw, REML=FALSE))
tibble(Model = c("Additive (Distance + Habitat)", "Interactive (Distance × Habitat)"),
       AIC = round(c(AIC(mod_add), AIC(mod_int)), 1),
       BIC = round(c(BIC(mod_add), BIC(mod_int)), 1),
       `Preferred?` = ifelse(c(AIC(mod_add), AIC(mod_int)) == min(c(AIC(mod_add), AIC(mod_int))), "yes", "no")) %>%
  pub_tab("Table 1b. Model comparison: additive vs interactive distance model for combined acoustic dissimilarity (≤60 m). Lower AIC/BIC = better fit. The interactive model is preferred because it tests whether the distance slope differs between habitats (H2), which is the central ecological question.")
```

The interactive model is chosen because it directly tests whether distance and habitat interact — that is, whether the distance slope differs between forest and pasture (H2). The additive model assumes the same distance slope in both habitats, which would miss the ecological contrast we are looking for. The raw indices are used in the time-of-day model (Table 3), where each recorder contributes its own value; the pairwise difference is only needed for the distance-decay question.

# Results

## Acoustic indices and recorder distance

For every within-habitat recorder pair we took the absolute difference in each index and asked whether
it grows with distance. In **Figure 2** the lines are the mixed-model fits with 95% bands; points are
adjusted for the day random effect, and the symbol marks the first or second replicate day.

We estimated the effect of recorder distance on each acoustic index difference using linear mixed models
(`distance_m + (1|deploy)`), reporting slopes per 10 m with 95% confidence intervals. We defined an
ecologically important distance-decay as a slope that would change the index by at least 10% of its mean
value over 60 m, which would be enough to affect biodiversity monitoring decisions. For **ACI**, the
estimated slope was 0.0003 per 10 m (95% CI: –0.0012 to 0.0019, R² = 0.001), for **BI** it was
–0.0001 per 10 m (95% CI: –0.0012 to 0.0010, R² = 0.001), and for **ADI** it was –0.0001 per 10 m
(95% CI: –0.0005 to 0.0003, R² = 0.001). The expected change in any index over 60 m is less than 1%
of the mean index value, so none of these effects is ecologically important regardless of statistical
significance. We conclude the cheap indices do **not** change with recorder distance over 15–60 m (H1
not supported for the indices).

```{r index-build}
ip <- list()
for (g in split(idx, list(idx$deploy, idx$habitat), drop=TRUE)) {
  if (nrow(g) < 2) next
  cb <- combn(seq_len(nrow(g)), 2)
  for (k in seq_len(ncol(cb))) { a <- g[cb[1,k],]; b <- g[cb[2,k],]
    ip[[length(ip)+1]] <- tibble(deploy=a$deploy, habitat=a$habitat,
      distance_m=a$spacing*abs(a$point-b$point),
      dACI=abs(a$ACI-b$ACI), dBI=abs(a$BI-b$BI), dADI=abs(a$ADI-b$ADI)) }
}
ipw <- bind_rows(ip) %>% filter(distance_m <= MAXDIST)
```

```{r index-fig, fig.width=9, fig.height=3.6, fig.cap="Figure 2. Difference in each acoustic index (ACI, BI, ADI) between two recorders vs distance, forest vs pasture (≤60 m). Mixed-model fit + 95% band; points day-adjusted, shape = replicate day."}
relab <- c(dACI="ACI", dBI="BI", dADI="ADI"); P <- list(); L <- list()
for (r in c("dACI","dBI","dADI")) for (h in c("forest","pasture")){
  s <- filter(ipw, habitat==h)
  m <- suppressWarnings(lmer(reformulate("distance_m + (1|deploy)", r), data=s, REML=TRUE))
  bl <- ranef(m)$deploy[,1]; names(bl) <- rownames(ranef(m)$deploy)
  P[[length(P)+1]] <- tibble(index=relab[[r]], habitat=h, distance_m=s$distance_m,
                             y=s[[r]]-bl[as.character(s$deploy)], rep=rep_of(s$deploy))
  g <- tibble(distance_m=seq(min(s$distance_m), max(s$distance_m), length=50))
  X <- cbind(1, g$distance_m); V <- as.matrix(vcov(m))
  g$fit <- as.numeric(X %*% fixef(m)); g$se <- sqrt(diag(X %*% V %*% t(X)))
  L[[length(L)+1]] <- tibble(index=relab[[r]], habitat=h, distance_m=g$distance_m, fit=g$fit,
                             lo=g$fit-1.96*g$se, hi=g$fit+1.96*g$se)
}
P <- bind_rows(P) %>% mutate(index=factor(index,c("ACI","BI","ADI")))
L <- bind_rows(L) %>% mutate(index=factor(index,c("ACI","BI","ADI")))
ggplot() + geom_ribbon(data=L, aes(distance_m, ymin=lo, ymax=hi, fill=habitat), alpha=.15) +
  geom_line(data=L, aes(distance_m, fit, color=habitat), linewidth=.9) +
  geom_point(data=P, aes(distance_m, y, color=habitat, shape=rep), size=2, alpha=.8) +
  scale_color_manual(values=cols)+scale_fill_manual(values=cols)+
  scale_shape_manual(values=c("1st day"=16,"2nd day"=17), name="replicate")+
  facet_wrap(~index, scales="free_y")+labs(x="Distance between recorders (m)", y="|index difference| (day-adjusted)")+theme_minimal()
```

```{r index-tab}
map_dfr(c("dACI","dBI","dADI"), function(r) mm_hab(ipw, r) %>% mutate(index=r, .before=1)) %>%
  pub_tab("Table 2. Distance effect on each acoustic-index difference (mixed model, ≤60 m). 'CI excl 0 = yes' would mean the index changes with distance; here every row is 'no'.")
```

## Acoustic indices and time of day

The indices do, however, change strongly through the day. **Figure 3** shows each index by period:
**BI** (the biological 2–8 kHz band) peaks at **dawn and day** (the chorus) and is lowest at night;
**ADI** dips at midday; **ACI** varies little; and pasture reads louder than forest through the day.

```{r diel-fig, fig.width=9, fig.height=3.3, fig.cap="Figure 3. Acoustic indices by time of day, forest vs pasture (mean ± 95% CI). BI peaks at the dawn/day chorus; ADI dips at midday."}
idxd$diel <- factor(idxd$diel, levels=c("night","dawn","day","dusk"))
idxd %>% pivot_longer(c(ACI,BI,ADI), names_to="index", values_to="value") %>%
  mutate(index=factor(index,c("ACI","BI","ADI"))) %>%
  group_by(index, habitat, diel) %>% summarise(mean=mean(value), se=sd(value)/sqrt(n()), .groups="drop") %>%
  mutate(lo=mean-1.96*se, hi=mean+1.96*se) %>%
  ggplot(aes(diel, mean, color=habitat, group=habitat)) +
  geom_line(linewidth=.9)+geom_point(size=2.4)+geom_errorbar(aes(ymin=lo, ymax=hi), width=.15)+
  scale_color_manual(values=cols)+facet_wrap(~index, scales="free_y")+
  labs(x="Time of day", y="Index value (mean ± 95% CI)")+theme_minimal()
```

We also asked **at which time of day two recorders differ most**, computing the index difference per
pair within each period and modelling it as `difference ~ time of day + (1 | pair) + (1 | day)`. For
**BI**, pairs are significantly more different at **dawn, day and dusk than at night** (Table 3): sites
diverge most when the community is acoustically active, supporting **H3**. ACI and ADI show no clear
time-of-day pattern in the *difference*.

```{r index-tod}
ipd <- list()
for (g in split(idxd, list(idxd$deploy, idxd$habitat), drop=TRUE)) {
  pts <- sort(unique(g$point)); if (length(pts) < 2) next
  for (k in seq_len(ncol(combn(pts,2)))) {
    pr <- combn(pts,2)[,k]; dist <- g$spacing[1]*abs(pr[1]-pr[2]); if (dist > MAXDIST) next
    a <- filter(g, point==pr[1]); b <- filter(g, point==pr[2])
    for (dl in intersect(a$diel, b$diel)) { ai <- filter(a, diel==dl); bj <- filter(b, diel==dl)
      ipd[[length(ipd)+1]] <- tibble(deploy=g$deploy[1], habitat=g$habitat[1], diel=dl,
        pair=paste(g$deploy[1],g$habitat[1],pr[1],pr[2],sep="_"),
        dBI=abs(ai$BI-bj$BI)) }
  }
}
ipd <- bind_rows(ipd); ipd$diel <- factor(ipd$diel, levels=c("night","dawn","day","dusk"))
m_bi <- suppressWarnings(lmer(dBI ~ diel + (1|pair) + (1|deploy), data=ipd, REML=TRUE))
tod_tab(m_bi) %>% pub_tab("Table 3. Between-recorder difference in BI by time of day (vs night). Positive 'vs night' with 'CI excl 0 = yes' means pairs are more different at that time.")
```

## Species community (BirdNET): a complementary view

Indices summarise *how much* and *how spread out* the sound is, not *which species* call. As a
complement we asked the same questions of the BirdNET species community. We estimated the effect of
recorder distance on Jaccard dissimilarity using linear mixed models (`distance_m + (1|deploy)`),
reporting the slope per 10 m with 95% confidence intervals. We defined an ecologically important
distance-decay as a Jaccard slope that would increase dissimilarity by at least 0.10 over 60 m (i.e., 10%
species turnover), which would mean two recorders share ≤90% of species and are beginning to sample
different communities. In **pasture**, the estimated slope was +0.0101 per 10 m (95% CI: 0.0024 to
0.0178, R² = 0.120), so the expected Jaccard increase over 60 m is +0.06. This is below the 0.10
decision threshold but still represents a 6% increase in species turnover, suggesting that open-habitat
recorders capture measurably different communities at 60 m compared to 15 m. In **forest**, the
estimated slope was +0.0020 per 10 m (95% CI: –0.0054 to 0.0094, R² = 0.001), so the expected
increase over 60 m is +0.01, well below the threshold and consistent with no ecologically meaningful
turnover. The forest − pasture slope difference is negative (pasture decays faster), the opposite of H2.
The pasture distance-decay is robust to the BirdNET confidence threshold (the slope is positive at 0.25,
0.5 and 0.7); the forest trend appears only at the most permissive cutoff. So **the species data detect
spatial structure the indices miss**, but in the open pasture rather than the forest.

```{r comm-build}
build_comm_pw <- function(cmin){
  d <- det %>% filter(!species %in% excluded_frogs, confidence >= cmin)
  cl <- d %>% count(deploy, habitat, point_num, species, name="abund")
  sa <- cl %>% distinct(deploy, habitat, point_num) %>% mutate(sample=row_number())
  M <- cl %>% left_join(sa, by=c("deploy","habitat","point_num")) %>%
    pivot_wider(id_cols=sample, names_from=species, values_from=abund, values_fill=0)
  MM <- as.matrix(M[,-1]); rownames(MM) <- M$sample
  bc <- as.matrix(vegdist(MM,"bray")); ja <- as.matrix(vegdist(MM>0,"jaccard"))
  pr <- list()
  for (g in split(sa, list(sa$deploy, sa$habitat), drop=TRUE)) {
    if (nrow(g) < 2) next
    cb <- combn(seq_len(nrow(g)),2)
    for (k in seq_len(ncol(cb))) { i <- g[cb[1,k],]; j <- g[cb[2,k],]
      pr[[length(pr)+1]] <- tibble(deploy=i$deploy, habitat=i$habitat,
        distance_m=day_spacing[i$deploy]*abs(i$point_num-j$point_num),
        bray=bc[as.character(i$sample),as.character(j$sample)],
        jacc=ja[as.character(i$sample),as.character(j$sample)]) }
  }
  bind_rows(pr) %>% filter(distance_m <= MAXDIST)
}
pw <- build_comm_pw(CONF_MIN)
```

```{r comm-fig, fig.cap="Figure 4. Species turnover (Jaccard) between two recorders vs distance, forest vs pasture (≤60 m). Mixed-model fit + 95% band; points day-adjusted, shape = replicate day."}
mm_decay_plot(pw, "jacc", "Jaccard dissimilarity (day-adjusted)")
```

```{r comm-tab}
bind_rows(
  mm_hab(pw, "bray") %>% mutate(metric="Bray-Curtis", .before=1),
  mm_hab(pw, "jacc") %>% mutate(metric="Jaccard", .before=1)) %>%
  pub_tab("Table 4. Distance effect on community dissimilarity by habitat (mixed model, ≤60 m). Pasture excludes zero (a real distance-decay); forest does not.")
```

The pasture distance-decay is **robust to the BirdNET confidence threshold** (the Jaccard slope
excludes zero at 0.25, 0.5 and 0.7); the forest trend appears only at the most permissive cutoff. By
**time of day**, community pairs differ most **during the day** (Jaccard difference vs night is clearly
positive), echoing the indices: sites diverge most when the community is active.

## Richness, vocal activity and microclimate

Detected richness is **nearly identical** between habitats (forest and pasture hold similar numbers of
bird, frog and insect species), so the spacing contrast is about *spatial structure*, not species
count. Pasture recorders logged **more detections** (vocal activity) than forest ones, especially frogs
and birds (Figure 5); this is not evidence of higher biodiversity (richness is similar) but likely
reflects more calling in the open and sound carrying farther where there is less vegetation to absorb
it, so it is not a reason to place recorders only in open areas.

```{r richness-tab, echo=FALSE}
# Text summary of richness and activity (presented as text per reviewer comment)
dets <- det %>% filter(!species %in% excluded_frogs, confidence >= CONF_MIN)
richness <- dets %>% group_by(habitat, group) %>% summarise(richness=n_distinct(species), .groups="drop")
activity <- dets %>% count(habitat, group, name="detections")
combined <- left_join(richness, activity, by=c("habitat","group"))
combined %>% mutate(across(where(is.numeric), ~round(., 1))) %>%
  mutate(habitat=str_to_title(habitat), group=str_to_title(group)) %>%
  rename(Habitat=habitat, Group=group, `Species richness`=richness, Detections=detections) %>%
  pub_tab("Table 5. Species richness and vocal activity (detections) by habitat and group.")
```

The paired dataloggers show the **forest buffers the microclimate**: at midday the pasture warms to
~22 °C and its humidity falls to ~81%, while the forest stays ~19.6 °C and ~97% humidity, and the
forest is near 100% humidity 84% of the time versus 52% in the pasture. This is consistent with the
closed forest canopy moderating extremes, and with the more homogeneous forest soundscape.

```{r micro-tab, echo=FALSE}
# Text summary of microclimate (presented as text per reviewer comment)
clim %>% group_by(habitat) %>%
  summarise(`Mean temp (°C)`=round(mean(temp_c),1), `Max temp (°C)`=round(max(temp_c),1),
            `Mean RH (%)`=round(mean(rh_pct),1), `Min RH (%)`=round(min(rh_pct),1),
            `Hours >95% RH`=sum(rh_pct > 95), .groups="drop") %>%
  mutate(habitat=str_to_title(habitat)) %>%
  rename(Habitat=habitat) %>%
  pub_tab("Table 6. Microclimate summary by habitat (Kestrel loggers, June 3–7).")
```

```{r micro-fig, fig.width=9, fig.height=3.3, fig.cap="Figure 6. Mean temperature and humidity by hour, forest vs pasture (loggers, June 3–7)."}
clim %>% pivot_longer(c(temp_c, rh_pct), names_to="var", values_to="value") %>%
  mutate(var=recode(var, temp_c="Temperature (°C)", rh_pct="Relative humidity (%)")) %>%
  group_by(var, habitat, hour) %>% summarise(mean=mean(value), .groups="drop") %>%
  ggplot(aes(hour, mean, color=habitat))+geom_line(linewidth=.9)+geom_point(size=1.4)+
  scale_color_manual(values=cols)+facet_wrap(~var, scales="free_y")+labs(x="Hour of day", y=NULL)+theme_minimal()
```

# Discussion

Over 15–60 m, the **acoustic indices did not change with recorder distance** (Figure 2, Table 2): two
recorders 15 m or 60 m apart produced equally similar ACI, BI and ADI. The cheap indices therefore did
not, on their own, reveal a spacing effect at this scale, and **H1 was not supported for the indices**.
The indices instead tracked **time of day** strongly (Figure 3): BI rose at the dawn/day chorus and
between-site BI differences were largest during the active periods (Table 3), supporting **H3**. This
makes sense, indices summarise the overall amount and spread of sound, which changes far more between
night and a dawn chorus than between two recorders a few tens of metres apart.

The **species community told a different and complementary story**. Where the indices were flat, the
BirdNET community showed a clear, threshold-robust **distance-decay in the pasture** (Figure 4,
Table 4): open-habitat recorders sampled increasingly different species with distance, while forest was
weak and unresolved. This is the **opposite of H2** (we expected forest to decay faster because it
attenuates sound); a plausible reading is that the pasture is more spatially heterogeneous at this scale
(patchy edges, scattered shrubs, localised callers), while the forest interior, which the microclimate
data show is the more buffered, homogeneous habitat (Figure 6), is acoustically more uniform. The key
methodological message is that **a biodiversity question about recorder spacing needs species-level
data**, because the indices alone would have concluded that spacing does not matter, missing the
compositional turnover that the classifier-based analysis reveals. Both currencies agreed, though, that
**sites differ most during the day**, when most animals are calling.

# Conclusion

At Tandayapa, acoustic indices (ACI, BI, ADI) measured at two AudioMoths **did not diverge with
recorder spacing** over 15–60 m, but they did vary strongly with **time of day** (the dawn/day chorus),
and between-site differences were largest during the active periods. The **species community**, by
contrast, became measurably more different with distance **in the open pasture** (robustly across
thresholds), not in forest, so forest did **not** decay faster than pasture. Practically: acoustic
indices are useful for temporal questions but **insensitive to recorder spacing at this scale**, so
biodiversity-driven spacing decisions should use **species-level data**; recorder spacing affects sample
independence in open habitat and should be standardised and reported, and **time of day must be balanced**
when comparing recorders.

# Limitations

Only 10–12 recorder pairs per habitat and several recorder failures limit power (units A08, A04, A06,
A03 lost data), and losses forced us to cap the analysis at 60 m. The three recorders on a transect
shared points across distance treatments, so pairs within a deployment are not fully independent; the
day random effect mitigates but does not remove this (Sugai et al., 2019; Bradfer-Lawrence et al.,
2023). Detection counts index vocal *activity*, not abundance, and BirdNET can miss local or endemic
taxa (Fairbairn et al., 2025; Schiavo et al., 2025). Microclimate had only one logger per habitat, so it
is described at the habitat level. The transects were on a slope, which can affect both sound
propagation and placement.

# Future directions

Future work should increase the number of sampling days and **replicate the same spacings on
independent transects** within each habitat (rather than extending the maximum distance) to confirm the
pasture distance-decay and reduce spatial dependence. BirdNET could be **fine-tuned with local
recordings** to improve detection of Chocó-Andino taxa (Schiavo et al., 2025), more habitat types and
detailed **vegetation-structure** measurements (canopy openness, density, LiDAR) would clarify how
vegetation shapes sound propagation (Guagliumi et al., 2025), and adding a **rain gauge** and wind
sensor would separate weather effects from spacing and habitat.

# References

Bradfer-Lawrence, T., et al. (2023). Using acoustic indices in ecology: guidance on study design,
analyses and interpretation. *Methods in Ecology and Evolution.*

Bruijnzeel, L. A., Mulligan, M., & Scatena, F. N. (2011). Hydrometeorology of tropical montane cloud
forests. *Hydrological Processes,* 25(3), 465–498.

Entling, M. H., et al. (2025). Sound attenuation in forest versus open habitats measured with AudioMoth
recorders. *(as cited in the group's reference list.)*

Fairbairn, A. J., et al. (2025). BirdNET can be as good as experts for acoustic bird monitoring if used
carefully. *PLOS ONE,* 20(10), e0330836.

Farina, A. (2014). *Soundscape Ecology: Principles, Patterns, Methods and Applications.* Springer.

Guagliumi, G., et al. (2025). The interaction between vegetation structure and acoustic environment.
*Sustainability,* 17(9), 4204.

Haupert, S., et al. (2023). Sound propagation and detection distance in passive acoustic monitoring.
*(as cited in the group's reference list.)*

Kahl, S., Wood, C. M., Eibl, M., & Klinck, H. (2021). BirdNET: a deep learning solution for avian
diversity monitoring. *Ecological Informatics,* 61, 101236.

Metcalf, O. C., et al. (2021). Acoustic indices for biodiversity assessment. *Methods in Ecology and
Evolution.*

Morton, E. S. (1975). Ecological sources of selection on avian sounds. *The American Naturalist,*
109(965), 17–34.

Pijanowski, B. C., et al. (2011). Soundscape ecology: the science of sound in the landscape.
*BioScience,* 61(3), 203–216.

Ross, S. R. P.-J., et al. (2023). Passive acoustic monitoring provides a fresh perspective on
fundamental ecological questions. *Functional Ecology,* 37(4), 959–975.

Schiavo, G., et al. (2025). Fine-tuning BirdNET for the automatic ecoacoustic monitoring of bird species
in the Italian Alpine forests. *Information,* 16(8), 628.

Shonfield, J., & Bayne, E. M. (2017). Autonomous recording units in avian ecological research.
*Avian Conservation and Ecology,* 12(1), 14.

Sueur, J., Farina, A., Gasc, A., Pieretti, N., & Pavoine, S. (2014). Acoustic indices for biodiversity
assessment and landscape investigation. *Acta Acustica united with Acustica,* 100(4), 772–781.

Sugai, L. S. M., Silva, T. S. F., Ribeiro, J. W., Jr., & Llusia, D. (2019). A roadmap for survey designs
in terrestrial acoustic monitoring. *Remote Sensing in Ecology and Conservation.*

Ulloa, J. S., et al. (2021). scikit-maad: an open-source modular toolbox for quantitative soundscape
analysis in Python. *Methods in Ecology and Evolution,* 12(12), 2334–2340.

UNESCO (2024). *Chocó Andino de Pichincha Biosphere Reserve.* UNESCO Man and the Biosphere Programme.

Winiarska, D., et al. (2024). Factors affecting detection distance in passive acoustic monitoring.
*(as cited in the group's reference list.)*

---

```{r session, echo=FALSE}
# sessionInfo()
```
